---
title: "AnchorMAGIC: does building the diffusion graph from a better feature space raise MAGIC's oracle?"
output: html_notebook
---

Every previous experiment tuned **t**. This one leaves the diffusion
identical and changes the **graph**. MAGIC's weakest assumption is that a
kNN graph built from PCA of noisy, dropout-heavy counts defines good
neighbourhoods. If the graph is wrong, no $t$ can fix it.

We build the cell--cell graph from several candidate "anchor" feature
spaces, diffuse the *same* full count matrix over each, and compare each
graph's **single-$t$ oracle** (best achievable TF$\rightarrow$target
recovery over $t\in\{0,\dots,15\}$).

**Why this is a fair test.** The graph is built with *no* TF labels
(fully unsupervised), and $t$ is oracle-selected identically for every
variant. So if an anchor graph's oracle exceeds the baseline graph's
oracle, the neighbourhood structure itself is genuinely better - a real
method improvement, not a selector or supervision artefact. (At $t=0$ all
graphs coincide with raw, so any difference must come from better
diffusion neighbourhoods.)

Graph variants:
\begin{itemize}
\item \textbf{base} - proper MAGIC graph: library-normalise, $\sqrt{}$,
  PCA of all panel genes (the standard preprocessing).
\item \textbf{pearson} - count-aware anchor: PCA of analytic Pearson
  residuals under a Poisson null (Townes 2019; Lause \& Kobak 2021),
  which denoises dropout before the kNN.
\item \textbf{hvg} - restrict to the top-200 most variable genes
  ($\sqrt{}$-PCA) - a crude "strong-signal anchor".
\item \textbf{module} - graph in gene-module-score space ($k$-means gene
  modules $\rightarrow$ per-cell mean scores).
\end{itemize}

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(cvMAGIC); library(Rmagic); library(Matrix); library(irlba); library(RANN)
  library(pROC); library(dplyr); library(tidyr); library(ggplot2); library(patchwork)
})
stopifnot(Rmagic::pymagic_is_available())
fig_dir  <- normalizePath(file.path("..","figures"), mustWork=FALSE)
data_dir <- normalizePath(file.path("..","data"),    mustWork=FALSE)
out_dir  <- normalizePath(file.path("..","notebooks"), mustWork=FALSE)
NPCA <- 20L; K <- 30L; T_GRID <- 0:15; SEEDS <- 101:103
```

```{r tf-lists}
tf_mouse <- list(
  Gata1=c("Klf1","Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1","Slc4a1","Gypa","Epor"),
  Gata2=c("Gata1","Klf1","Itga2b"), Klf1=c("Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1","Slc4a1"),
  Sfpi1=c("Cebpa","Mpo","Csf1r","Csf2ra","Csf2rb"), Cebpa=c("Mpo","Elane","Prtn3","Ctsg","Csf3r"),
  Cebpe=c("Mpo","Elane","Prtn3","Ctsg"), Gfi1=c("Csf1r","Elane"), Irf8=c("Cxcr4"),
  Tal1=c("Gata1","Klf1","Gata2"))
tf_tusi <- tf_mouse; names(tf_tusi)[names(tf_tusi)=="Sfpi1"]<-"Spi1"
tf_tusi <- lapply(tf_tusi, function(v){v[v=="Hbb-b1"]<-"Hbb-bs"; v[v=="Hbb-b2"]<-"Hbb-bt"; v})
tf_human <- list(
  PAX5=c("CD19","MS4A1","CD79A","CD79B","BLNK","BANK1","BCL11A"),
  TBX21=c("IFNG","GZMB","CXCR3","GNLY","PRF1","NKG7"), EOMES=c("IFNG","GZMA","PRF1","GNLY","NKG7","KLRD1"),
  FOXP3=c("IL2RA","CTLA4","TNFRSF18"), SPI1=c("CD14","LYZ","CSF1R","FCGR3A","CST3","S100A8","S100A9"),
  IRF8=c("CCR7","FLT3","CST3","HLA-DRA"), CEBPB=c("CD14","LYZ","S100A8","S100A9"),
  LEF1=c("SELL","CCR7","TCF7"), KLF2=c("SELL","S1PR1","CCR7"), ID3=c("SELL","CCR7"))
```

```{r helpers}
build_panel <- function(M, tf, min_lib, gene_pos, seed, n_cells=1500L, n_genes=2000L) {
  ls<-rowSums(M); keep<-which(ls>=min_lib & ls<=quantile(ls,0.99))
  set.seed(seed); ci<-sample(keep,min(n_cells,length(keep))); cn<-M[ci,,drop=FALSE]
  mk<-intersect(unique(c(names(tf),unlist(tf))),colnames(cn))
  gp<-colMeans(cn>0); cn<-cn[,gp>=gene_pos|colnames(cn)%in%mk,drop=FALSE]
  gv<-apply(cn,2,var); hv<-order(gv,decreasing=TRUE)[seq_len(min(n_genes,length(gv)))]
  X<-cn[,union(colnames(cn)[hv],mk),drop=FALSE]; storage.mode(X)<-"integer"; X
}
fast_auc <- function(scores, labels){ r<-rank(scores); np<-sum(labels==1); nn<-sum(labels==0)
  if(np==0||nn==0) return(NA_real_); (sum(r[labels==1])-np*(np+1)/2)/(np*nn) }
make_pairs <- function(tf, gs, seed=7L){ set.seed(seed); rows<-list()
  for(f in names(tf)){ if(!(f%in%gs))next; tg<-intersect(tf[[f]],gs); if(!length(tg))next
    pool<-setdiff(gs,c(f,tg)); ng<-sample(pool,min(length(pool),5L*length(tg)))
    rows[[f]]<-rbind(data.frame(tf=f,gene=tg,label=1L),data.frame(tf=f,gene=ng,label=0L)) }
  do.call(rbind,rows) }

# ---- embeddings (cells x d) ----
norm_sqrt <- function(X){ s<-rowSums(X); s[s==0]<-1; Xn<-X/s*mean(s); sqrt(Xn) }
emb_base <- function(X) irlba::prcomp_irlba(norm_sqrt(X), n=NPCA, center=TRUE)$x
emb_hvg  <- function(X,n=200L){ v<-apply(X,2,var); top<-order(v,decreasing=TRUE)[seq_len(min(n,ncol(X)))]
  irlba::prcomp_irlba(norm_sqrt(X[,top,drop=FALSE]), n=min(NPCA,length(top)-1L), center=TRUE)$x }
emb_pearson <- function(X){                      # analytic Pearson residuals, Poisson null
  X<-as.matrix(X); n<-rowSums(X); p<-colSums(X)/sum(X); mu<-outer(n,p)
  r<-(X-mu)/sqrt(mu+1e-8); cl<-sqrt(nrow(X)); r[r>cl]<-cl; r[r< -cl]<- -cl
  irlba::prcomp_irlba(r, n=NPCA, center=TRUE)$x }
emb_module <- function(X,M=25L){ Xs<-norm_sqrt(X)
  gp<-scale(t(Xs)); npc<-min(15L,nrow(gp)-1L,ncol(gp)-1L)
  pcg<-irlba::prcomp_irlba(gp,n=npc,center=TRUE)$x
  set.seed(1L); km<-stats::kmeans(pcg,centers=M,nstart=5L)
  sc<-sapply(seq_len(M),function(m) rowMeans(Xs[,km$cluster==m,drop=FALSE]))
  scale(sc) }

# ---- graph from a precomputed embedding (mirrors magic_graph kernel) ----
graph_from_emb <- function(P,k=K,decay=1){
  N<-nrow(P); nn<-RANN::nn2(P,P,k=k+1L); idx<-nn$nn.idx[,-1,drop=FALSE]; dst<-nn$nn.dists[,-1,drop=FALSE]
  sig<-dst[,k]; sig[sig==0]<-.Machine$double.eps; aff<-exp(-(dst/sig)^decay)
  A<-Matrix::sparseMatrix(i=rep.int(seq_len(N),times=k),j=as.integer(idx),x=as.numeric(aff),dims=c(N,N))
  A<-(A+Matrix::t(A))/2; rs<-Matrix::rowSums(A); rs[rs==0]<-1; A/rs }

# AUC-vs-t curve for a given graph M, diffusing full raw X, scoring pairs
curve_for_graph <- function(Mmat, X, pairs){
  gs<-colnames(X); used<-unique(c(pairs$tf,pairs$gene))
  Y<-X; prev<-0L; aucs<-numeric(length(T_GRID))
  for(i in seq_along(T_GRID)){ t<-T_GRID[i]
    while(prev<t){ Y<-as.matrix(Mmat%*%Y); prev<-prev+1L }
    imp<-if(t==0L) as.matrix(X) else Y; colnames(imp)<-gs
    C<-suppressWarnings(cor(imp[,used,drop=FALSE],method="spearman"))
    sc<-abs(mapply(function(a,b) C[a,b],pairs$tf,pairs$gene)); sc[is.na(sc)]<-0
    aucs[i]<-fast_auc(sc,pairs$label) }
  aucs }
```

```{r loaders}
load_paul<-function(){cf<-data.table::fread(file.path(data_dir,"GSE72857_umitab.txt"),sep="\t",header=TRUE,data.table=FALSE);gn<-cf[[1]];cf<-as.matrix(cf[,-1]);rownames(cf)<-gn;M<-t(cf);storage.mode(M)<-"integer";M}
load_pbmc<-function(){d<-file.path(data_dir,"filtered_gene_bc_matrices","hg19");m<-Matrix::readMM(file.path(d,"matrix.mtx"));gt<-read.table(file.path(d,"genes.tsv"),sep="\t");rownames(m)<-make.unique(gt$V2);m<-as.matrix(t(m));storage.mode(m)<-"integer";m}
load_tm<-function(){d<-file.path(data_dir,"droplet","Marrow-10X_P7_2");m<-Matrix::readMM(file.path(d,"matrix.mtx"));gt<-read.table(file.path(d,"genes.tsv"),sep="\t");rownames(m)<-make.unique(gt$V1);m<-as.matrix(t(m));storage.mode(m)<-"integer";m}
load_tusi<-function(){csv<-file.path(data_dir,"GSM2388072_basal_bone_marrow.raw_umifm_counts.csv");df<-data.table::fread(csv,sep=",",header=TRUE,data.table=FALSE);meta<-c("cell_id","barcode","library_id","seq_run_id","pass_filter");M<-as.matrix(df[,setdiff(colnames(df),meta)]);rownames(M)<-df$cell_id;storage.mode(M)<-"integer";M}
specs<-list(
  Paul=list(load=load_paul,tf=tf_mouse,min_lib=500L,gene_pos=0.05),
  PBMC=list(load=load_pbmc,tf=tf_human,min_lib=500L,gene_pos=0.03),
  TM  =list(load=load_tm,  tf=tf_mouse,min_lib=500L,gene_pos=0.03),
  Tusi=list(load=load_tusi,tf=tf_tusi, min_lib=1000L,gene_pos=0.05))
EMB<-list(base=emb_base, pearson=emb_pearson, hvg=emb_hvg, module=emb_module)
```

```{r run, message=FALSE}
rows<-list(); curves<-list()
for(nm in names(specs)){
  sp<-specs[[nm]]; M<-sp$load()
  cat(sprintf("\n=== %s ===\n",nm))
  for(s in SEEDS){
    X<-build_panel(M,sp$tf,sp$min_lib,sp$gene_pos,seed=s); gs<-colnames(X); pr<-make_pairs(sp$tf,gs)
    rmg<-as.matrix(Rmagic::magic(X,t="auto",knn=K,npca=NPCA,verbose=0)$result); colnames(rmg)<-gs
    C<-suppressWarnings(cor(rmg[,unique(c(pr$tf,pr$gene)),drop=FALSE],method="spearman"))
    scr<-abs(mapply(function(a,b)C[a,b],pr$tf,pr$gene)); scr[is.na(scr)]<-0; auc_rmagic<-fast_auc(scr,pr$label)
    for(en in names(EMB)){
      P<-EMB[[en]](X); Mmat<-graph_from_emb(P)
      crv<-curve_for_graph(Mmat,X,pr); to<-T_GRID[which.max(crv)]
      rows[[paste(nm,s,en)]]<-data.frame(dataset=nm,seed=s,graph=en,
        oracle_t=to, oracle_auc=max(crv), raw_auc=crv[1], rmagic_auc=auc_rmagic)
      if(s==SEEDS[1]) curves[[paste(nm,en)]]<-data.frame(dataset=nm,graph=en,t=T_GRID,auc=crv)
    }
    base_or<-rows[[paste(nm,s,"base")]]$oracle_auc
    cat(sprintf("  seed %d | base=%.3f pearson=%.3f hvg=%.3f module=%.3f | rmagic=%.3f\n",s,
      base_or, rows[[paste(nm,s,"pearson")]]$oracle_auc, rows[[paste(nm,s,"hvg")]]$oracle_auc,
      rows[[paste(nm,s,"module")]]$oracle_auc, auc_rmagic))
  }
  rm(M); invisible(gc())
}
res<-do.call(rbind,rows); curve_df<-do.call(rbind,curves)
write.csv(res,file.path(out_dir,"anchor_magic_results.csv"),row.names=FALSE)
write.csv(curve_df,file.path(out_dir,"anchor_magic_curves.csv"),row.names=FALSE)
```

```{r summary}
agg<-res|>group_by(dataset,graph)|>summarise(oracle=mean(oracle_auc),oracle_t=median(oracle_t),.groups="drop")|>
  pivot_wider(names_from=graph,values_from=c(oracle,oracle_t))
base<-res|>filter(graph=="base")|>group_by(dataset)|>summarise(base=mean(oracle_auc),.groups="drop")
delta<-res|>group_by(dataset,graph)|>summarise(oracle=mean(oracle_auc),.groups="drop")|>
  left_join(base,by="dataset")|>mutate(vs_base=oracle-base)|>
  select(dataset,graph,oracle,vs_base)|>arrange(match(dataset,c("Paul","PBMC","TM","Tusi")),graph)
cat("Oracle TF-target AUC per graph (mean over",length(SEEDS),"seeds), and anchor - base:\n")
print(as.data.frame(delta),row.names=FALSE,digits=3)
cat("\nDoes any anchor graph beat the BASE graph's oracle? (per dataset, mean over seeds)\n")
for(d in unique(res$dataset)){ b<-delta$oracle[delta$dataset==d & delta$graph=="base"]
  for(g in c("pearson","hvg","module")){ v<-delta$vs_base[delta$dataset==d & delta$graph==g]
    cat(sprintf("  %-5s %-8s %+.4f%s\n",d,g,v,ifelse(v>0," (better)",""))) } }
```

```{r fig, fig.width=11, fig.height=4}
curve_df$dataset<-factor(curve_df$dataset,levels=c("Paul","PBMC","TM","Tusi"))
pal<-c(base="#1f77b4",pearson="#2ca02c",hvg="#9467bd",module="#ff7f0e")
fig<-ggplot(curve_df,aes(t,auc,colour=graph))+geom_line(linewidth=.7)+geom_point(size=1)+
  facet_wrap(~dataset,nrow=1,scales="free_y")+scale_colour_manual(values=pal,name="graph")+
  scale_x_continuous(breaks=seq(0,15,3))+
  labs(x="diffusion time t",y="TF-target AUC (full pairs)",
       title="AnchorMAGIC: recovery vs t under different graph constructions (seed 101)")+
  theme_classic(base_size=10)+theme(legend.position="bottom")
ggsave(file.path(fig_dir,"fig_anchor_magic.png"),fig,width=11,height=4,dpi=220,bg="white")
fig
```
