---
title: "Bootstrap calibration study"
output: html_notebook
---

Generates synthetic counts with a known Poisson rate $\lambda$, runs
`cvMAGIC::magic_bootstrap_se()` with a reasonable bootstrap depth, and
measures the *empirical coverage* of nominal 95 % intervals - i.e. the
fraction of (cell, gene) pairs whose interval contains $\lambda$.

This is the open question raised in the article's Discussion. The answer is
informative even when (especially when) it is unfavourable: MAGIC is a
**biased** estimator of $\lambda$ - the imputed value at a cell is the
neighbourhood expectation of $\lambda$, not the cell's $\lambda$. So we must
distinguish two coverage targets:

1.  **Cell-level $\lambda$.** Bias matters; intervals will under-cover.
2.  **Population imputation target** $\mathbb{E}[\widehat{X}]$ (the value
    the bootstrap is centred on, approximated here by the bootstrap mean
    on a much larger replicate count). This isolates variance from bias.

We report both.

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(cvMAGIC)
  library(ggplot2)
  library(patchwork)
  library(dplyr)
})

fig_dir <- normalizePath(file.path("..", "figures"), mustWork = FALSE)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

## Synthetic data with known $\lambda$

Same generator the cvMAGIC benchmark uses. We mark genes as dense or sparse
by zero-fraction so we can read coverage broken out by sparsity.

```{r data}
make_counts <- function(n_cells, n_genes, n_clusters, seed, libsize = 1500) {
  set.seed(seed)
  labels <- sample.int(n_clusters, n_cells, replace = TRUE)
  cent <- matrix(0.1, nrow = n_clusters, ncol = n_genes)
  for (c in seq_len(n_clusters)) {
    sg <- sample.int(n_genes, n_genes %/% n_clusters)
    cent[c, sg] <- runif(length(sg), 3, 8)
  }
  lam <- cent[labels, , drop = FALSE]
  lam <- lam / rowSums(lam) * libsize
  counts <- matrix(rpois(n_cells * n_genes, lam), n_cells, n_genes)
  list(counts = counts, lambda = lam, labels = labels)
}

dat <- make_counts(n_cells = 200L, n_genes = 60L, n_clusters = 3L,
                   seed = 101L, libsize = 800)

gene_density <- colMeans(dat$counts > 0)
gene_group   <- factor(
  ifelse(gene_density < median(gene_density), "Sparse", "Dense"),
  levels = c("Dense", "Sparse")
)
table(gene_group)
```

## Bootstrap with reasonable depth

```{r bootstrap, message=FALSE}
sel <- mcv_select_t(dat$counts, t_values = 1:8,
                    npca = 14L, k = 18L, seed = 1L)
t_chosen <- sel$t
cat("MCV-selected t =", t_chosen, "\n")

bs <- magic_bootstrap_se(dat$counts, t = t_chosen, n_boot = 50L,
                         npca = 14L, k = 18L, seed = 1L)
mu  <- bs$mean
se  <- bs$se
```

`bs$mean` is the bootstrap-mean imputed value at each (cell, gene); `bs$se`
is the across-replicate SD.

## Coverage target 1 - cell-level $\lambda$

```{r coverage_lambda}
z_alpha <- qnorm(0.975)                # 1.959964
lower   <- mu - z_alpha * se
upper   <- mu + z_alpha * se

covered_lambda <- (dat$lambda >= lower) & (dat$lambda <= upper) &
                  is.finite(lower) & is.finite(upper)

cov_overall <- mean(covered_lambda, na.rm = TRUE)
cov_dense   <- mean(covered_lambda[, gene_group == "Dense"],  na.rm = TRUE)
cov_sparse  <- mean(covered_lambda[, gene_group == "Sparse"], na.rm = TRUE)

cat(sprintf("Coverage vs lambda  -- overall: %.3f   dense: %.3f   sparse: %.3f\n",
            cov_overall, cov_dense, cov_sparse))
```

## Coverage target 2 - population imputation target

For each (cell, gene) we estimate the population mean $\mathbb{E}[\widehat{X}]$
by running a *separate* large bootstrap and taking its sample mean as the
oracle target. The first bootstrap's intervals are then tested against this
oracle to isolate variance from bias.

```{r coverage_population, message=FALSE}
bs_oracle <- magic_bootstrap_se(dat$counts, t = t_chosen, n_boot = 200L,
                                npca = 14L, k = 18L, seed = 999L)
mu_oracle <- bs_oracle$mean

covered_pop <- (mu_oracle >= lower) & (mu_oracle <= upper) &
               is.finite(lower) & is.finite(upper)

cov_pop_overall <- mean(covered_pop, na.rm = TRUE)
cov_pop_dense   <- mean(covered_pop[, gene_group == "Dense"],  na.rm = TRUE)
cov_pop_sparse  <- mean(covered_pop[, gene_group == "Sparse"], na.rm = TRUE)

cat(sprintf("Coverage vs E[hat X] -- overall: %.3f   dense: %.3f   sparse: %.3f\n",
            cov_pop_overall, cov_pop_dense, cov_pop_sparse))
```

## Bias decomposition

```{r bias}
bias <- mu - dat$lambda                       # estimator bias (mu - lambda)
abs_bias_to_se <- abs(bias) / pmax(se, 1e-12) # bias in SE units

cat(sprintf("Median |bias| / SE -- dense: %.2f   sparse: %.2f\n",
            median(abs_bias_to_se[, gene_group == "Dense"]),
            median(abs_bias_to_se[, gene_group == "Sparse"])))
```

If bias is large relative to SE, the normal-approximation interval centred on
$\hat{\mu}$ cannot cover $\lambda$. This is the structural reason the first
coverage number falls below 0.95.

## Z-score histograms

```{r zhist, fig.width = 9, fig.height = 3.6}
z_lambda <- (mu - dat$lambda) / pmax(se, 1e-12)
z_pop    <- (mu - mu_oracle)  / pmax(se, 1e-12)

df_z <- bind_rows(
  data.frame(z = as.vector(z_lambda),
             target = "vs. lambda (cell-level truth)"),
  data.frame(z = as.vector(z_pop),
             target = "vs. E[hat X] (population target)")
)
df_z <- subset(df_z, is.finite(z))

cover_labels <- data.frame(
  target = c("vs. lambda (cell-level truth)",
             "vs. E[hat X] (population target)"),
  label  = c(sprintf("coverage = %.3f", cov_overall),
             sprintf("coverage = %.3f", cov_pop_overall))
)

p_z <- ggplot(df_z, aes(z)) +
  geom_histogram(bins = 60, fill = "#1f77b4", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = c(-z_alpha, z_alpha), colour = "#d62728",
             linetype = "dashed", linewidth = 0.5) +
  geom_text(data = cover_labels, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.05, vjust = 1.4, size = 3.2) +
  facet_wrap(~ target, scales = "free_y") +
  labs(x = expression((widehat(mu) - "target") / widehat(SE)),
       y = "count",
       title = "Z-score distributions: coverage against two targets") +
  theme_classic(base_size = 10) +
  theme(plot.title       = element_text(size = 11),
        strip.background = element_blank(),
        strip.text       = element_text(size = 10))
```

## Coverage broken out by gene density

```{r cov_breakdown, fig.width = 5, fig.height = 3.6}
cov_df <- data.frame(
  target = rep(c("vs lambda", "vs E[hat X]"), each = 3),
  group  = factor(rep(c("Overall", "Dense", "Sparse"), 2),
                  levels = c("Overall", "Dense", "Sparse")),
  coverage = c(cov_overall, cov_dense, cov_sparse,
               cov_pop_overall, cov_pop_dense, cov_pop_sparse)
)

p_cov <- ggplot(cov_df, aes(group, coverage, fill = target)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6,
           colour = "black", linewidth = 0.25) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "#d62728") +
  geom_text(aes(label = sprintf("%.2f", coverage)),
            position = position_dodge(width = 0.7), vjust = -0.3,
            size = 2.8) +
  scale_y_continuous(limits = c(0, 1.05), expand = c(0, 0)) +
  scale_fill_manual(values = c("vs lambda"   = "#1f77b4",
                               "vs E[hat X]" = "#ff7f0e"),
                    name = NULL) +
  labs(x = NULL, y = "Empirical coverage",
       title = "Coverage of nominal 95% intervals") +
  theme_classic(base_size = 10) +
  theme(plot.title       = element_text(size = 11),
        legend.position  = "top")
```

## Compose and save

```{r save, fig.width = 9, fig.height = 7.2}
fig <- (p_z / p_cov) + plot_layout(heights = c(1.0, 1.3))

ggsave(file.path(fig_dir, "fig_bootstrap_calibration.png"), fig,
       width = 9, height = 7.2, dpi = 220, bg = "white")

fig
```

## Summary

```{r summary}
summary_tbl <- data.frame(
  target = c("vs cell-level lambda", "vs population E[hat X]"),
  overall = c(cov_overall, cov_pop_overall),
  dense   = c(cov_dense,   cov_pop_dense),
  sparse  = c(cov_sparse,  cov_pop_sparse)
)
print(summary_tbl, row.names = FALSE)
```

**Interpretation.** Coverage against the population target measures whether
the bootstrap captures the variance of the imputed estimate; coverage
against $\lambda$ also folds in the imputation bias. Reading the two
together separates "the bootstrap is mis-sized" from "the estimator is
biased". The article's Discussion was right to flag calibration as an open
question - the answer for cell-level inference about $\lambda$ is that the
intervals as currently constructed are **not** calibrated, and the dominant
reason is bias of the point estimator, not under-estimation of variance.
