---
title: "msMAGIC: a learned multi-scale diffusion mixture - does it beat the best single t?"
output: html_notebook
---

The paper shows no *unsupervised* $t$-selector reliably beats MAGIC.
This notebook tests a different class of method: instead of one diffusion
time, learn a **convex mixture of diffusion scales**
\[
  X_{\mathrm{msMAGIC}} = \sum_{t=0}^{T} w_t\,M^t X,
  \qquad w_t \ge 0,\ \textstyle\sum_t w_t = 1,
\]
a non-negative polynomial graph filter $p(M)X$. Weights are learned to
maximise held-out TF$\rightarrow$target recovery on **train pairs**, and
every number is reported on disjoint **test pairs** (pair-level CV).

**Fairness.** The mixture's search space contains every single-$t$
imputation (the simplex vertices), so it can only beat a single $t$ on
*train*. The honest question is whether it beats, on **held-out test
pairs**, the single-$t$ oracle that is *also chosen on train*. That is
the decisive comparison; `rmagic_auto`, `cv_mcv`, and `raw` are
references, and the test-selected single-$t$ is the ceiling.

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(cvMAGIC); library(Rmagic); library(Matrix); library(dplyr); library(tidyr)
  library(ggplot2); library(patchwork)
})
stopifnot(Rmagic::pymagic_is_available())
fig_dir  <- normalizePath(file.path("..", "figures"), mustWork = FALSE)
data_dir <- normalizePath(file.path("..", "data"),    mustWork = FALSE)
out_dir  <- normalizePath(file.path("..", "notebooks"), mustWork = FALSE)
NPCA <- 20L; K <- 30L; T_SET <- c(0L,1L,2L,3L,4L,6L,8L,10L); CELL_SEED <- 101L; NFOLD <- 5L
```

```{r tf-lists}
tf_mouse <- list(
  Gata1=c("Klf1","Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1","Slc4a1","Gypa","Epor"),
  Gata2=c("Gata1","Klf1","Itga2b"), Klf1=c("Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1","Slc4a1"),
  Sfpi1=c("Cebpa","Mpo","Csf1r","Csf2ra","Csf2rb"), Cebpa=c("Mpo","Elane","Prtn3","Ctsg","Csf3r"),
  Cebpe=c("Mpo","Elane","Prtn3","Ctsg"), Gfi1=c("Csf1r","Elane"), Irf8=c("Cxcr4"),
  Tal1=c("Gata1","Klf1","Gata2"))
tf_tusi <- tf_mouse; names(tf_tusi)[names(tf_tusi)=="Sfpi1"] <- "Spi1"
tf_tusi <- lapply(tf_tusi, function(v){v[v=="Hbb-b1"]<-"Hbb-bs"; v[v=="Hbb-b2"]<-"Hbb-bt"; v})
tf_human <- list(
  PAX5=c("CD19","MS4A1","CD79A","CD79B","BLNK","BANK1","BCL11A"),
  TBX21=c("IFNG","GZMB","CXCR3","GNLY","PRF1","NKG7"), EOMES=c("IFNG","GZMA","PRF1","GNLY","NKG7","KLRD1"),
  FOXP3=c("IL2RA","CTLA4","TNFRSF18"), SPI1=c("CD14","LYZ","CSF1R","FCGR3A","CST3","S100A8","S100A9"),
  IRF8=c("CCR7","FLT3","CST3","HLA-DRA"), CEBPB=c("CD14","LYZ","S100A8","S100A9"),
  LEF1=c("SELL","CCR7","TCF7"), KLF2=c("SELL","S1PR1","CCR7"), ID3=c("SELL","CCR7"))
```

```{r helpers}
build_panel <- function(M, tf, min_lib, gene_pos, seed, n_cells=1500L, n_genes=2000L) {
  ls <- rowSums(M); keep <- which(ls>=min_lib & ls<=quantile(ls,0.99))
  set.seed(seed); ci <- sample(keep, min(n_cells,length(keep))); cn <- M[ci,,drop=FALSE]
  mk <- intersect(unique(c(names(tf),unlist(tf))), colnames(cn))
  gp <- colMeans(cn>0); cn <- cn[, gp>=gene_pos | colnames(cn)%in%mk, drop=FALSE]
  gv <- apply(cn,2,var); hv <- order(gv,decreasing=TRUE)[seq_len(min(n_genes,length(gv)))]
  X <- cn[, union(colnames(cn)[hv], mk), drop=FALSE]; storage.mode(X)<-"integer"; X
}
# Mann-Whitney AUC from scores + 0/1 labels
fast_auc <- function(scores, labels) {
  r <- rank(scores); np <- sum(labels==1); nn <- sum(labels==0)
  if (np==0 || nn==0) return(NA_real_)
  (sum(r[labels==1]) - np*(np+1)/2) / (np*nn)
}
# Build the pair table for a panel: positives = curated targets, negatives = 5x random non-targets
make_pairs <- function(tf, gene_set, seed=7L) {
  set.seed(seed); rows <- list()
  for (f in names(tf)) {
    if (!(f %in% gene_set)) next
    targs <- intersect(tf[[f]], gene_set); if (!length(targs)) next
    pool <- setdiff(gene_set, c(f, targs))
    negs <- sample(pool, min(length(pool), 5L*length(targs)))
    rows[[f]] <- rbind(
      data.frame(tf=f, gene=targs, label=1L),
      data.frame(tf=f, gene=negs,  label=0L))
  }
  do.call(rbind, rows)
}
softmax <- function(th){ e<-exp(th-max(th)); e/sum(e) }
```

```{r core}
# Given a panel X, precompute the diffusion basis and the per-method results.
eval_msmagic <- function(X, tf, label) {
  gs <- colnames(X); g <- magic_graph(X, npca=NPCA, k=K)
  # diffusion basis Y_t (cells x genes) for t in T_SET
  basis <- list(); Y <- X; prev <- 0L
  for (t in T_SET) { while (prev < t) { Y <- as.matrix(g$M %*% Y); prev <- prev+1L }
    basis[[as.character(t)]] <- if (t==0L) as.matrix(X) else Y }
  # rmagic + cv_mcv (unsupervised references)
  rmg <- as.matrix(Rmagic::magic(X, t="auto", knn=K, npca=NPCA, verbose=0)$result); colnames(rmg)<-gs
  sel <- mcv_select_t(X, t_values=0:10, npca=NPCA, k=K, seed=1L)
  mcv_imp <- if (sel$t==0L) as.matrix(X) else magic_impute(X, t=sel$t, graph=g); colnames(mcv_imp)<-gs

  pairs <- make_pairs(tf, gs)
  used <- unique(c(pairs$tf, pairs$gene))
  basis_sub <- lapply(basis, function(M) M[, used, drop=FALSE])  # 1500 x n_used per t
  # precompute spearman-rank scoring for a fixed imputed matrix over a pair subset
  score_auc <- function(imp_used, pr, method=c("spearman","pearson")) {
    method <- match.arg(method)
    C <- suppressWarnings(cor(imp_used, method=method))
    sc <- abs(mapply(function(a,b) C[a,b], pr$tf, pr$gene)); sc[is.na(sc)] <- 0
    fast_auc(sc, pr$label)
  }
  imp_used <- function(M) M[, used, drop=FALSE]
  mix_used <- function(w) Reduce(`+`, Map(function(M,wt) M*wt, basis_sub, w))
  # Pearson-margin surrogate (smooth, fast) for optimisation
  margin_obj <- function(th, pr) {
    w <- softmax(th); MM <- mix_used(w)
    C <- suppressWarnings(cor(MM, method="pearson"))
    sc <- abs(mapply(function(a,b) C[a,b], pr$tf, pr$gene)); sc[is.na(sc)] <- 0
    -(mean(sc[pr$label==1]) - mean(sc[pr$label==0]))   # maximise margin
  }
  learn_w <- function(train) {
    nT <- length(T_SET)
    starts <- c(list(rep(0,nT), rep(0,nT)),                 # uniform x2
                lapply(seq_len(nT), function(i){v<-rep(-3,nT);v[i]<-3;v}))  # each single-t vertex
    best <- NULL; bestval <- Inf
    for (st in starts) {
      o <- tryCatch(optim(st, margin_obj, pr=train, method="Nelder-Mead",
                          control=list(maxit=250)), error=function(e) NULL)
      if (!is.null(o) && o$value < bestval) { bestval <- o$value; best <- o$par }
    }
    softmax(best)
  }

  # pair-level CV
  set.seed(11L); fold <- sample(rep_len(seq_len(NFOLD), nrow(pairs)))
  rmg_u <- imp_used(rmg); mcv_u <- imp_used(mcv_imp); raw_u <- basis_sub[["0"]]
  basis_auc_test <- function(pr) vapply(T_SET, function(t)
      score_auc(basis_sub[[as.character(t)]], pr), numeric(1))
  res <- list()
  for (f in seq_len(NFOLD)) {
    tr <- pairs[fold!=f,]; te <- pairs[fold==f,]
    # single-t selected on TRAIN, judged on TEST
    tr_by_t <- basis_auc_test(tr); t_star <- T_SET[which.max(tr_by_t)]
    auc_single_test <- score_auc(basis_sub[[as.character(t_star)]], te)
    # test-selected single-t (ceiling)
    te_by_t <- basis_auc_test(te); auc_oracle_test <- max(te_by_t)
    # msMAGIC: learn w on train, judge on test
    w <- learn_w(tr); MM <- mix_used(w)
    auc_ms_test <- score_auc(MM, te)
    auc_ms_train <- score_auc(MM, tr)
    res[[f]] <- data.frame(
      dataset=label, fold=f, t_single=t_star,
      raw=score_auc(raw_u,te), rmagic=score_auc(rmg_u,te), cv_mcv=score_auc(mcv_u,te),
      single_t=auc_single_test, msMAGIC=auc_ms_test, msMAGIC_train=auc_ms_train,
      oracle_test=auc_oracle_test, w0=w[1], wmax_t=T_SET[which.max(w)])
  }
  list(res=do.call(rbind,res), mcv_t=sel$t)
}
```

```{r loaders}
load_paul <- function(){ cf<-data.table::fread(file.path(data_dir,"GSE72857_umitab.txt"),sep="\t",header=TRUE,data.table=FALSE); gn<-cf[[1]]; cf<-as.matrix(cf[,-1]); rownames(cf)<-gn; M<-t(cf); storage.mode(M)<-"integer"; M }
load_pbmc <- function(){ d<-file.path(data_dir,"filtered_gene_bc_matrices","hg19"); m<-Matrix::readMM(file.path(d,"matrix.mtx")); gt<-read.table(file.path(d,"genes.tsv"),sep="\t"); rownames(m)<-make.unique(gt$V2); m<-as.matrix(t(m)); storage.mode(m)<-"integer"; m }
load_tm <- function(){ d<-file.path(data_dir,"droplet","Marrow-10X_P7_2"); m<-Matrix::readMM(file.path(d,"matrix.mtx")); gt<-read.table(file.path(d,"genes.tsv"),sep="\t"); rownames(m)<-make.unique(gt$V1); m<-as.matrix(t(m)); storage.mode(m)<-"integer"; m }
load_tusi <- function(){ csv<-file.path(data_dir,"GSM2388072_basal_bone_marrow.raw_umifm_counts.csv"); df<-data.table::fread(csv,sep=",",header=TRUE,data.table=FALSE); meta<-c("cell_id","barcode","library_id","seq_run_id","pass_filter"); M<-as.matrix(df[,setdiff(colnames(df),meta)]); rownames(M)<-df$cell_id; storage.mode(M)<-"integer"; M }
specs <- list(
  Paul=list(load=load_paul, tf=tf_mouse, min_lib=500L, gene_pos=0.05),
  PBMC=list(load=load_pbmc, tf=tf_human, min_lib=500L, gene_pos=0.03),
  TM  =list(load=load_tm,   tf=tf_mouse, min_lib=500L, gene_pos=0.03),
  Tusi=list(load=load_tusi, tf=tf_tusi,  min_lib=1000L, gene_pos=0.05))
```

```{r run, message=FALSE}
all_res <- list()
for (nm in names(specs)) {
  sp <- specs[[nm]]; M <- sp$load()
  X <- build_panel(M, sp$tf, sp$min_lib, sp$gene_pos, seed=CELL_SEED)
  cat(sprintf("\n=== %s : %d x %d ===\n", nm, nrow(X), ncol(X)))
  e <- eval_msmagic(X, sp$tf, nm); all_res[[nm]] <- e$res
  r <- e$res
  cat(sprintf("  test AUC means | raw=%.3f rmagic=%.3f cv_mcv=%.3f single_t=%.3f msMAGIC=%.3f oracle=%.3f (mcv t=%d, median w-peak t=%g, mean w0=%.2f)\n",
              mean(r$raw),mean(r$rmagic),mean(r$cv_mcv),mean(r$single_t),
              mean(r$msMAGIC),mean(r$oracle_test), e$mcv_t,
              median(r$wmax_t), mean(r$w0)))
  rm(M); invisible(gc())
}
res <- do.call(rbind, all_res)
write.csv(res, file.path(out_dir,"msmagic_results.csv"), row.names=FALSE)
```

```{r summary}
agg <- res |> group_by(dataset) |>
  summarise(raw=mean(raw), rmagic=mean(rmagic), cv_mcv=mean(cv_mcv),
            single_t=mean(single_t), msMAGIC=mean(msMAGIC), oracle=mean(oracle_test),
            ms_vs_single=mean(msMAGIC-single_t), ms_vs_rmagic=mean(msMAGIC-rmagic),
            ms_train=mean(msMAGIC_train), .groups="drop") |>
  arrange(match(dataset, c("Paul","PBMC","TM","Tusi")))
cat("Held-out TEST Spearman AUC (mean over", NFOLD, "pair folds):\n")
print(as.data.frame(agg), row.names=FALSE, digits=3)

# paired test: does msMAGIC beat single-t oracle on held-out pairs?
cat("\nmsMAGIC - single_t_oracle (per dataset, paired over folds):\n")
for (d in unique(res$dataset)) {
  x <- res$msMAGIC[res$dataset==d] - res$single_t[res$dataset==d]
  cat(sprintf("  %-5s mean=%+.4f  >0 in %d/%d folds\n", d, mean(x), sum(x>0), length(x)))
}
```

```{r fig, fig.width=10, fig.height=4}
long <- agg |> select(dataset, raw, rmagic, cv_mcv, single_t, msMAGIC, oracle) |>
  pivot_longer(-dataset, names_to="method", values_to="AUC") |>
  mutate(method=factor(method, levels=c("raw","rmagic","cv_mcv","single_t","msMAGIC","oracle")),
         dataset=factor(dataset, levels=c("Paul","PBMC","TM","Tusi")))
pal <- c(raw="grey70", rmagic="#ff7f0e", cv_mcv="#1f77b4", single_t="#9467bd",
         msMAGIC="#2ca02c", oracle="#d62728")
fig <- ggplot(long, aes(method, AUC, fill=method)) +
  geom_col(width=.7, colour="black", linewidth=.2) +
  facet_wrap(~dataset, nrow=1, scales="free_y") +
  scale_fill_manual(values=pal, name=NULL) +
  labs(x=NULL, y="held-out TEST TF-target AUC",
       title="msMAGIC (learned multi-scale mixture) vs single-t oracle vs upstream - all on held-out pairs") +
  theme_classic(base_size=10) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), legend.position="bottom")
ggsave(file.path(fig_dir,"fig_msmagic.png"), fig, width=10, height=4, dpi=220, bg="white")
fig
```
