---
title: "cvMAGIC paper, consolidated four-dataset benchmark"
output: html_notebook
---

This is the single canonical harness behind the paper's main results. It
runs **one identical evaluation** (`eval_all`) over all four datasets -
Paul 2015 (MARS-seq), 10x PBMC 3k, Tabula Muris marrow (10x droplet),
and Tusi 2018 (inDrop), across multiple cell-subsample seeds, and emits:

* `paper_benchmark_results.csv`: one row per (dataset, seed) with the
  chosen t and AUC for raw / rmagic_auto / vanilla MCV / tolerant MCV /
  oracle.
* `paper_benchmark_curves.csv`: the seed-1 AUC-vs-t curve per dataset.
* `figures/fig_paper_summary.png`: cross-dataset head-to-head.
* `figures/fig_paper_divergence.png`: AUC-vs-t with each selector's
  landing vs the oracle (the prediction-vs-recovery divergence figure).

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(cvMAGIC); library(Rmagic); library(Matrix); library(pROC)
  library(ggplot2); library(dplyr); library(tidyr); library(patchwork)
})
stopifnot(Rmagic::pymagic_is_available())
fig_dir  <- normalizePath(file.path("..", "figures"), mustWork = FALSE)
data_dir <- normalizePath(file.path("..", "data"),    mustWork = FALSE)
out_dir  <- normalizePath(file.path("..", "notebooks"), mustWork = FALSE)
NPCA <- 20L; K <- 30L; SEEDS <- 101:105; T_SEL <- 0:10; T_ORACLE <- 0:15
```

```{r tf-lists}
# Mouse hematopoietic programme (Paul symbol convention)
tf_mouse <- list(
  Gata1 = c("Klf1","Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1","Slc4a1","Gypa","Epor"),
  Gata2 = c("Gata1","Klf1","Itga2b"),
  Klf1  = c("Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1","Slc4a1"),
  Sfpi1 = c("Cebpa","Mpo","Csf1r","Csf2ra","Csf2rb"),
  Cebpa = c("Mpo","Elane","Prtn3","Ctsg","Csf3r"),
  Cebpe = c("Mpo","Elane","Prtn3","Ctsg"),
  Gfi1  = c("Csf1r","Elane"),
  Irf8  = c("Cxcr4"),
  Tal1  = c("Gata1","Klf1","Gata2"))
# Tusi inDrop symbol variants: Sfpi1->Spi1, Hbb-b1/b2 -> Hbb-bs/bt
tf_tusi <- tf_mouse
names(tf_tusi)[names(tf_tusi) == "Sfpi1"] <- "Spi1"
tf_tusi <- lapply(tf_tusi, function(v) {
  v[v == "Hbb-b1"] <- "Hbb-bs"; v[v == "Hbb-b2"] <- "Hbb-bt"; v })
# Human immune programme (PBMC)
tf_human <- list(
  PAX5  = c("CD19","MS4A1","CD79A","CD79B","BLNK","BANK1","BCL11A"),
  TBX21 = c("IFNG","GZMB","CXCR3","GNLY","PRF1","NKG7"),
  EOMES = c("IFNG","GZMA","PRF1","GNLY","NKG7","KLRD1"),
  FOXP3 = c("IL2RA","CTLA4","TNFRSF18"),
  SPI1  = c("CD14","LYZ","CSF1R","FCGR3A","CST3","S100A8","S100A9"),
  IRF8  = c("CCR7","FLT3","CST3","HLA-DRA"),
  CEBPB = c("CD14","LYZ","S100A8","S100A9"),
  LEF1  = c("SELL","CCR7","TCF7"),
  KLF2  = c("SELL","S1PR1","CCR7"),
  ID3   = c("SELL","CCR7"))
```

```{r harness}
score_method <- function(imp_mat, tf_targets, gene_set) {
  pos <- list(); neg <- list(); set.seed(7L)
  for (tf in names(tf_targets)) {
    if (!(tf %in% gene_set)) next
    targets <- intersect(tf_targets[[tf]], gene_set)
    if (length(targets) == 0L) next
    pool <- setdiff(gene_set, c(tf, targets))
    non_targets <- sample(pool, min(length(pool), 5L * length(targets)))
    tf_vec <- imp_mat[, tf]
    cp <- abs(apply(imp_mat[, targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cn <- abs(apply(imp_mat[, non_targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cp[is.na(cp)] <- 0; cn[is.na(cn)] <- 0
    pos[[tf]] <- data.frame(score = cp, label = 1L)
    neg[[tf]] <- data.frame(score = cn, label = 0L)
  }
  pairs <- rbind(do.call(rbind, pos), do.call(rbind, neg))
  list(auc = as.numeric(pROC::roc(pairs$label, pairs$score, levels = c(0L,1L),
         direction = "<", quiet = TRUE)$auc),
       n_pos = sum(pairs$label == 1L), n_neg = sum(pairs$label == 0L))
}

build_panel <- function(M, tf_targets, min_lib, gene_pos, seed,
                        n_cells = 1500L, n_genes = 2000L) {
  libsize <- rowSums(M)
  keep <- which(libsize >= min_lib & libsize <= quantile(libsize, 0.99))
  set.seed(seed); ci <- sample(keep, min(n_cells, length(keep)))
  cn <- M[ci, , drop = FALSE]
  must_keep <- intersect(unique(c(names(tf_targets), unlist(tf_targets))),
                         colnames(cn))
  gpos <- colMeans(cn > 0)
  cn <- cn[, gpos >= gene_pos | colnames(cn) %in% must_keep, drop = FALSE]
  gv <- apply(cn, 2, var)
  hv <- order(gv, decreasing = TRUE)[seq_len(min(n_genes, length(gv)))]
  X <- cn[, union(colnames(cn)[hv], must_keep), drop = FALSE]
  storage.mode(X) <- "integer"; X
}

eval_all <- function(X, tf_targets) {
  gs <- colnames(X); g <- magic_graph(X, npca = NPCA, k = K)
  auc_at <- function(t) {
    imp <- if (t == 0L) X else magic_impute(X, t = t, graph = g)
    colnames(imp) <- gs; score_method(imp, tf_targets, gs)$auc
  }
  curve <- vapply(T_ORACLE, auc_at, numeric(1)); names(curve) <- T_ORACLE
  sel <- mcv_select_t(X, t_values = T_SEL, npca = NPCA, k = K, seed = 1L)
  t_tol <- sel$t_values[max(which(sel$loss <= min(sel$loss) * 1.05))]
  rmg <- as.matrix(Rmagic::magic(X, t = "auto", knn = K, npca = NPCA,
                                 verbose = 0)$result); colnames(rmg) <- gs
  list(curve = curve, row = data.frame(
    t_vanilla = sel$t, auc_vanilla = curve[as.character(sel$t)],
    t_tol = t_tol,     auc_tol     = curve[as.character(t_tol)],
    t_oracle = as.integer(names(which.max(curve))), auc_oracle = max(curve),
    auc_rmagic = score_method(rmg, tf_targets, gs)$auc,
    auc_raw = curve["0"], row.names = NULL))
}
```

```{r loaders}
load_paul <- function() {
  cf <- data.table::fread(file.path(data_dir, "GSE72857_umitab.txt"),
                          sep = "\t", header = TRUE, data.table = FALSE)
  gn <- cf[[1]]; cf <- as.matrix(cf[, -1]); rownames(cf) <- gn
  M <- t(cf); storage.mode(M) <- "integer"; M
}
load_pbmc <- function() {
  d <- file.path(data_dir, "filtered_gene_bc_matrices", "hg19")
  mat <- Matrix::readMM(file.path(d, "matrix.mtx"))
  gtsv <- read.table(file.path(d, "genes.tsv"), sep = "\t", header = FALSE)
  rownames(mat) <- make.unique(gtsv$V2)
  mat <- as.matrix(t(mat)); storage.mode(mat) <- "integer"; mat
}
load_tm <- function() {
  d <- file.path(data_dir, "droplet", "Marrow-10X_P7_2")
  mat <- Matrix::readMM(file.path(d, "matrix.mtx"))
  gtsv <- read.table(file.path(d, "genes.tsv"), sep = "\t", header = FALSE)
  rownames(mat) <- make.unique(gtsv$V1)
  mat <- as.matrix(t(mat)); storage.mode(mat) <- "integer"; mat
}
load_tusi <- function() {
  csv <- file.path(data_dir, "GSM2388072_basal_bone_marrow.raw_umifm_counts.csv")
  df <- data.table::fread(csv, sep = ",", header = TRUE, data.table = FALSE)
  meta <- c("cell_id","barcode","library_id","seq_run_id","pass_filter")
  M <- as.matrix(df[, setdiff(colnames(df), meta)]); rownames(M) <- df$cell_id
  storage.mode(M) <- "integer"; M
}

specs <- list(
  Paul = list(load = load_paul, tf = tf_mouse, min_lib = 500L,  gene_pos = 0.05,
              platform = "MARS-seq", structure = "trajectory"),
  PBMC = list(load = load_pbmc, tf = tf_human, min_lib = 500L,  gene_pos = 0.03,
              platform = "10x",      structure = "discrete"),
  TM   = list(load = load_tm,   tf = tf_mouse, min_lib = 500L,  gene_pos = 0.03,
              platform = "10x",      structure = "discrete"),
  Tusi = list(load = load_tusi, tf = tf_tusi,  min_lib = 1000L, gene_pos = 0.05,
              platform = "inDrop",   structure = "trajectory"))
```

```{r run, message=FALSE}
rows <- list(); curves <- list()
for (nm in names(specs)) {
  sp <- specs[[nm]]
  M <- sp$load()
  cat(sprintf("\n=== %s (%s, %s) : %d cells x %d genes loaded ===\n",
              nm, sp$platform, sp$structure, nrow(M), ncol(M)))
  for (s in SEEDS) {
    X <- build_panel(M, sp$tf, sp$min_lib, sp$gene_pos, seed = s)
    e <- eval_all(X, sp$tf)
    r <- e$row; r$dataset <- nm; r$seed <- s
    r$platform <- sp$platform; r$structure <- sp$structure
    r$median_umi <- median(rowSums(X))
    rows[[paste(nm, s)]] <- r
    if (s == SEEDS[1]) curves[[nm]] <- data.frame(
      dataset = nm, t = as.integer(names(e$curve)), AUC = e$curve,
      t_vanilla = r$t_vanilla, t_tol = r$t_tol, t_oracle = r$t_oracle)
    cat(sprintf("  seed %d | van t=%d(%.3f) tol t=%d(%.3f) oracle t=%d(%.3f) rmagic=%.3f raw=%.3f\n",
                s, r$t_vanilla, r$auc_vanilla, r$t_tol, r$auc_tol,
                r$t_oracle, r$auc_oracle, r$auc_rmagic, r$auc_raw))
  }
  rm(M); invisible(gc())
}
res <- do.call(rbind, rows)
curve_df <- do.call(rbind, curves)
write.csv(res, file.path(out_dir, "paper_benchmark_results.csv"), row.names = FALSE)
write.csv(curve_df, file.path(out_dir, "paper_benchmark_curves.csv"), row.names = FALSE)
```

```{r summary-table}
ord <- c("Paul","PBMC","TM","Tusi")
agg <- res |> group_by(dataset) |>
  summarise(platform = first(platform), structure = first(structure),
            t_oracle = median(t_oracle), t_vanilla = median(t_vanilla),
            t_tol = median(t_tol),
            raw = mean(auc_raw), rmagic = mean(auc_rmagic),
            mcv = mean(auc_vanilla), tol = mean(auc_tol),
            oracle = mean(auc_oracle),
            mcv_vs_rmagic = mean(auc_vanilla - auc_rmagic),
            tol_vs_mcv = mean(auc_tol - auc_vanilla), .groups = "drop") |>
  arrange(match(dataset, ord))
cat("Four-dataset summary (means over", length(SEEDS), "seeds):\n")
print(as.data.frame(agg), row.names = FALSE, digits = 3)
```

```{r fig-summary, fig.width = 11, fig.height = 4.5}
long <- res |>
  pivot_longer(c(auc_raw, auc_rmagic, auc_vanilla, auc_tol, auc_oracle),
               names_to = "method", values_to = "AUC") |>
  mutate(method = recode(method, auc_raw = "raw", auc_rmagic = "rmagic_auto",
           auc_vanilla = "cv_mcv", auc_tol = "magic_tolerant", auc_oracle = "oracle"),
         method = factor(method, levels = c("raw","rmagic_auto","cv_mcv",
           "magic_tolerant","oracle")),
         dataset = factor(dataset, levels = ord))
sumv <- long |> group_by(dataset, method) |>
  summarise(m = mean(AUC), sd = sd(AUC), .groups = "drop")
pal <- c(raw = "grey70", rmagic_auto = "#ff7f0e", cv_mcv = "#1f77b4",
         magic_tolerant = "#2ca02c", oracle = "#d62728")
fig_sum <- ggplot(sumv, aes(method, m, fill = method)) +
  geom_col(width = .7, colour = "black", linewidth = .2) +
  geom_errorbar(aes(ymin = m - sd, ymax = m + sd), width = .25) +
  facet_wrap(~ dataset, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = pal, name = NULL) +
  labs(x = NULL, y = "TF-target ROC AUC",
       title = "cvMAGIC across four datasets: MCV t-selection vs upstream MAGIC vs oracle") +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "bottom")
ggsave(file.path(fig_dir, "fig_paper_summary.png"), fig_sum,
       width = 11, height = 4.5, dpi = 220, bg = "white")
fig_sum
```

```{r fig-divergence, fig.width = 11, fig.height = 4}
curve_df$dataset <- factor(curve_df$dataset, levels = ord)
marks <- curve_df |> group_by(dataset) |>
  summarise(tv = first(t_vanilla), tt = first(t_tol), to = first(t_oracle),
            .groups = "drop")
pts <- bind_rows(
  curve_df |> inner_join(marks |> transmute(dataset, t = tv), by = c("dataset","t")) |>
    mutate(sel = "MCV / tolerant pick"),
  curve_df |> inner_join(marks |> transmute(dataset, t = to), by = c("dataset","t")) |>
    mutate(sel = "oracle"))
fig_div <- ggplot(curve_df, aes(t, AUC)) +
  geom_line(colour = "grey45", linewidth = .7) +
  geom_point(colour = "grey45", size = 1.3) +
  geom_point(data = pts, aes(colour = sel), size = 3) +
  facet_wrap(~ dataset, nrow = 1, scales = "free_y") +
  scale_colour_manual(values = c("MCV / tolerant pick" = "#1f77b4",
                                 "oracle" = "#d62728"), name = NULL) +
  scale_x_continuous(breaks = seq(0, 15, 3)) +
  labs(x = "diffusion time t", y = "TF-target AUC",
       title = "Prediction-vs-recovery divergence: where MCV lands vs the AUC oracle") +
  theme_classic(base_size = 10) + theme(legend.position = "bottom")
ggsave(file.path(fig_dir, "fig_paper_divergence.png"), fig_div,
       width = 11, height = 4, dpi = 220, bg = "white")
fig_div
```
