---
title: "Paul et al. 2015 bone-marrow benchmark"
output: html_notebook
---

Real-data follow-up requested in the article's Discussion: compare
`rmagic_auto`, `cv_mcv`, and `cv_per_module` on the Paul et al. 2015
mouse bone-marrow dataset (GEO **GSE72857**), the dataset used in the
original MAGIC paper's downstream validation.

We report two of the three metrics named in the Discussion:

1. **Cluster preservation.** Adjusted Rand Index between Louvain
   clusters on the imputed matrix and Louvain clusters on the raw data.
2. **Transcription-factor target prediction AUC.** For each method,
   treat known hematopoietic TF-target pairs as positives and random
   TF-non-target pairs as negatives; score each pair by the magnitude
   of its Spearman correlation in the imputed expression matrix;
   compute ROC AUC.

The third metric named in the article, FACS-protein correlation, is
**not applicable**: Paul 2015 used a MARS-seq protocol with no paired
FACS protein measurements per cell.

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(cvMAGIC)
  library(Rmagic)
  library(Matrix)
  library(Seurat)
  library(mclust)              # adjustedRandIndex
  library(pROC)                # ROC AUC
  library(ggplot2)
  library(dplyr)
  library(patchwork)
})
stopifnot(Rmagic::pymagic_is_available())

fig_dir <- normalizePath(file.path("..", "figures"), mustWork = FALSE)
data_dir <- normalizePath(file.path("..", "data"), mustWork = FALSE)
dir.create(fig_dir,  showWarnings = FALSE, recursive = TRUE)
dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
```

## Download (once)

```{r download, message=FALSE}
url      <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE72nnn/GSE72857/suppl/GSE72857_umitab.txt.gz"
gz_path  <- file.path(data_dir, "GSE72857_umitab.txt.gz")
txt_path <- file.path(data_dir, "GSE72857_umitab.txt")

if (!file.exists(gz_path)) {
  options(timeout = 600)
  utils::download.file(url, gz_path, mode = "wb", quiet = TRUE)
}
if (!file.exists(txt_path)) {
  R.utils::gunzip(gz_path, destname = txt_path, remove = FALSE)
}

cat("File size on disk:", file.info(txt_path)$size, "bytes\n")
```

## Load, subsample, filter

The full table is ~27 k genes x ~10 k cells. We subsample cells and pick
high-variance genes to keep the notebook tractable. The subset is large
enough that the methods can differ; running on the full table is a
matter of changing `n_cells_keep` and `n_genes_keep`.

```{r load, message=FALSE}
counts_full <- data.table::fread(txt_path, sep = "\t",
                                 header = TRUE, data.table = FALSE)
gene_names <- counts_full[[1]]
counts_full <- as.matrix(counts_full[, -1, drop = FALSE])
rownames(counts_full) <- gene_names
dim(counts_full)                  # genes x cells

# Drop cells with very low or very high library size.
libsize <- colSums(counts_full)
keep_cell <- libsize >= 500 & libsize <= quantile(libsize, 0.99)
cat("Cells passing libsize filter:", sum(keep_cell), "/", length(keep_cell), "\n")

# Subsample cells.
set.seed(101L)
n_cells_keep <- 1500L
cell_idx <- sample(which(keep_cell), min(n_cells_keep, sum(keep_cell)))
cnts <- counts_full[, cell_idx, drop = FALSE]

# Filter to genes expressed in >= 5% of cells, then top high-variance.
gene_pos_frac <- rowMeans(cnts > 0)
cnts <- cnts[gene_pos_frac >= 0.05, , drop = FALSE]
gene_var <- apply(cnts, 1, var)
n_genes_keep <- 2000L
gene_idx <- order(gene_var, decreasing = TRUE)[seq_len(
  min(n_genes_keep, length(gene_var)))]
cnts <- cnts[gene_idx, , drop = FALSE]

# Transpose to cells x genes (cvMAGIC convention).
X <- t(cnts)
storage.mode(X) <- "integer"
dim(X)
```

## Run the three methods

```{r impute, message=FALSE}
NPCA <- 20L; K <- 30L

t_start <- Sys.time()

# cv_mcv: MCV-tuned t.
sel_mcv <- mcv_select_t(X, t_values = 1:10,
                        npca = NPCA, k = K, seed = 1L)
g_full  <- magic_graph(X, npca = NPCA, k = K)
imp_cv_mcv <- magic_impute(X, t = sel_mcv$t, graph = g_full)
cat(sprintf("cv_mcv         t = %d  (took %.1fs)\n",
            sel_mcv$t,
            as.numeric(difftime(Sys.time(), t_start, units = "secs"))))

# cv_per_module: per-module adaptive t.
t_step <- Sys.time()
out_modules <- magic_modules(X, n_modules = 8L, t_values = 1:10,
                             npca = NPCA, k = K, seed = 1L)
imp_cv_per_module <- out_modules$imputed
cat(sprintf("cv_per_module  t = %s  (took %.1fs)\n",
            paste(out_modules$t_per_module, collapse = ","),
            as.numeric(difftime(Sys.time(), t_step, units = "secs"))))

# rmagic_auto: upstream Python MAGIC with t = "auto".
t_step <- Sys.time()
X_named <- X
colnames(X_named) <- rownames(cnts)
rownames(X_named) <- colnames(cnts)
out_rmagic <- Rmagic::magic(X_named, t = "auto", knn = K, npca = NPCA,
                            verbose = 0)
imp_rmagic_auto <- as.matrix(out_rmagic$result)
cat(sprintf("rmagic_auto    t = auto  (took %.1fs)\n",
            as.numeric(difftime(Sys.time(), t_step, units = "secs"))))

# Sanity: imputed matrices should have the same dims as X.
stopifnot(all(dim(imp_cv_mcv)        == dim(X)))
stopifnot(all(dim(imp_cv_per_module) == dim(X)))
stopifnot(all(dim(imp_rmagic_auto)   == dim(X)))
```

## Cluster preservation

We cluster each matrix (raw and three imputed) with Seurat's
graph-based Louvain at a common resolution, then measure ARI of the
imputed clusterings against the raw clustering. "Preservation" here
means: does imputation keep the cluster structure visible in the raw
data, or distort it?

```{r cluster}
cluster_louvain <- function(mat, npca = 20L, k_param = 20L, res = 0.6) {
  rownames(mat) <- paste0("cell", seq_len(nrow(mat)))
  colnames(mat) <- paste0("g",   seq_len(ncol(mat)))
  obj <- CreateSeuratObject(counts = t(mat), assay = "RNA")
  obj <- SetAssayData(obj, assay = "RNA", layer = "data",
                      new.data = log1p(t(mat)))
  VariableFeatures(obj) <- colnames(mat)
  obj <- ScaleData(obj, verbose = FALSE)
  obj <- RunPCA(obj, npcs = npca, verbose = FALSE,
                features = VariableFeatures(obj))
  obj <- FindNeighbors(obj, dims = seq_len(npca), k.param = k_param,
                       verbose = FALSE)
  obj <- FindClusters(obj, resolution = res, verbose = FALSE,
                      random.seed = 1L)
  as.integer(as.character(Idents(obj)))
}

cl_raw <- cluster_louvain(X)
cl_mcv <- cluster_louvain(imp_cv_mcv)
cl_per <- cluster_louvain(imp_cv_per_module)
cl_rma <- cluster_louvain(imp_rmagic_auto)

cluster_tbl <- data.frame(
  method   = c("cv_mcv", "cv_per_module", "rmagic_auto"),
  n_clust  = c(length(unique(cl_mcv)),
               length(unique(cl_per)),
               length(unique(cl_rma))),
  ari_raw  = c(adjustedRandIndex(cl_mcv, cl_raw),
               adjustedRandIndex(cl_per, cl_raw),
               adjustedRandIndex(cl_rma, cl_raw))
)
cat(sprintf("Raw clustering: %d clusters\n", length(unique(cl_raw))))
print(cluster_tbl, row.names = FALSE)
```

## TF-target prediction AUC

Curated hematopoietic TF-target relationships (from the lineage-marker
literature: Tusi et al. 2018, Olsson et al. 2016, Paul et al. 2015).
The list is intentionally small (high-confidence) so the comparison is
between methods, not between databases.

```{r tf-targets}
tf_targets <- list(
  Gata1 = c("Klf1", "Hba-a1", "Hba-a2", "Hbb-b1", "Hbb-b2",
            "Aqp1", "Slc4a1", "Gypa", "Epor"),
  Gata2 = c("Gata1", "Klf1", "Itga2b"),
  Klf1  = c("Hba-a1", "Hba-a2", "Hbb-b1", "Hbb-b2", "Aqp1", "Slc4a1"),
  Sfpi1 = c("Cebpa", "Mpo", "Csf1r", "Csf2ra", "Csf2rb"),
  Cebpa = c("Mpo", "Elane", "Prtn3", "Ctsg", "Csf3r"),
  Cebpe = c("Mpo", "Elane", "Prtn3", "Ctsg"),
  Gfi1  = c("Csf1r", "Elane"),
  Irf8  = c("Cxcr4"),
  Tal1  = c("Gata1", "Klf1", "Gata2")
)

gene_set <- colnames(X_named)
score_method <- function(imp_mat) {
  pos <- data.frame(tf = character(), target = character(),
                    score = numeric(), label = integer())
  neg <- data.frame(tf = character(), target = character(),
                    score = numeric(), label = integer())
  set.seed(7L)
  for (tf in names(tf_targets)) {
    if (!(tf %in% gene_set)) next
    targets <- intersect(tf_targets[[tf]], gene_set)
    if (length(targets) == 0L) next
    # Skip the TF itself from the non-target pool.
    non_target_pool <- setdiff(gene_set, c(tf, targets))
    n_neg <- min(length(non_target_pool), 5L * length(targets))
    non_targets <- sample(non_target_pool, n_neg)

    tf_vec <- imp_mat[, tf]
    # |Spearman| (rank correlation, robust to MAGIC's smoothing).
    cor_pos <- abs(suppressWarnings(
      apply(imp_mat[, targets, drop = FALSE], 2, function(g)
        cor(tf_vec, g, method = "spearman"))))
    cor_neg <- abs(suppressWarnings(
      apply(imp_mat[, non_targets, drop = FALSE], 2, function(g)
        cor(tf_vec, g, method = "spearman"))))
    cor_pos[is.na(cor_pos)] <- 0
    cor_neg[is.na(cor_neg)] <- 0

    pos <- rbind(pos, data.frame(tf = tf, target = targets,
                                 score = cor_pos, label = 1L))
    neg <- rbind(neg, data.frame(tf = tf, target = non_targets,
                                 score = cor_neg, label = 0L))
  }
  all_pairs <- rbind(pos, neg)
  auc <- as.numeric(pROC::roc(all_pairs$label, all_pairs$score,
                              levels = c(0L, 1L), direction = "<",
                              quiet = TRUE)$auc)
  list(auc = auc, pairs = all_pairs)
}

# Add column names back to imputed matrices.
colnames(imp_cv_mcv)        <- colnames(X_named)
colnames(imp_cv_per_module) <- colnames(X_named)
colnames(imp_rmagic_auto)   <- colnames(X_named)

raw_named <- X_named
auc_raw <- score_method(raw_named)
auc_mcv <- score_method(imp_cv_mcv)
auc_per <- score_method(imp_cv_per_module)
auc_rma <- score_method(imp_rmagic_auto)

cat(sprintf("\nTF-target AUC\n"))
cat(sprintf("  raw            %.3f  (%d pos pairs scored)\n",
            auc_raw$auc, sum(auc_raw$pairs$label == 1)))
cat(sprintf("  cv_mcv         %.3f\n", auc_mcv$auc))
cat(sprintf("  cv_per_module  %.3f\n", auc_per$auc))
cat(sprintf("  rmagic_auto    %.3f\n", auc_rma$auc))
```

## Summary table

```{r summary}
summary_tbl <- data.frame(
  method       = c("raw",   "cv_mcv",        "cv_per_module", "rmagic_auto"),
  n_clusters   = c(length(unique(cl_raw)), length(unique(cl_mcv)),
                   length(unique(cl_per)), length(unique(cl_rma))),
  ARI_vs_raw   = c(1.0, cluster_tbl$ari_raw),
  TF_target_AUC = c(auc_raw$auc, auc_mcv$auc, auc_per$auc, auc_rma$auc)
)
print(summary_tbl, row.names = FALSE)
```

## Figure

```{r plot, fig.width = 8.5, fig.height = 3.4}
plot_df <- summary_tbl[summary_tbl$method != "raw", ]
plot_df$method <- factor(plot_df$method,
                         levels = c("cv_mcv", "cv_per_module", "rmagic_auto"))

palette <- c("cv_mcv" = "#1f77b4",
             "cv_per_module" = "#5fa8d3",
             "rmagic_auto"   = "#ff7f0e")

p_ari <- ggplot(plot_df, aes(method, ARI_vs_raw, fill = method)) +
  geom_col(width = 0.6, colour = "black", linewidth = 0.25) +
  geom_text(aes(label = sprintf("%.3f", ARI_vs_raw)),
            vjust = -0.4, size = 3) +
  scale_fill_manual(values = palette, guide = "none") +
  scale_y_continuous(limits = c(0, max(plot_df$ARI_vs_raw) * 1.18),
                     expand = c(0, 0)) +
  labs(x = NULL, y = "ARI (vs. raw clustering)",
       title = "(a) Cluster preservation") +
  theme_classic(base_size = 10) +
  theme(plot.title = element_text(size = 11),
        axis.text.x = element_text(angle = 25, hjust = 1))

# Include `raw` in the AUC bar to show the lift from imputation.
auc_df <- summary_tbl
auc_df$method <- factor(auc_df$method,
                        levels = c("raw", "cv_mcv", "cv_per_module",
                                   "rmagic_auto"))
auc_palette <- c("raw" = "#bbbbbb", palette)

p_auc <- ggplot(auc_df, aes(method, TF_target_AUC, fill = method)) +
  geom_col(width = 0.6, colour = "black", linewidth = 0.25) +
  geom_text(aes(label = sprintf("%.3f", TF_target_AUC)),
            vjust = -0.4, size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  scale_fill_manual(values = auc_palette, guide = "none") +
  scale_y_continuous(limits = c(0, max(auc_df$TF_target_AUC) * 1.18),
                     expand = c(0, 0)) +
  labs(x = NULL, y = "ROC AUC",
       title = "(b) TF-target prediction") +
  theme_classic(base_size = 10) +
  theme(plot.title = element_text(size = 11),
        axis.text.x = element_text(angle = 25, hjust = 1))

fig <- p_ari + p_auc + plot_layout(widths = c(1, 1.25))

ggsave(file.path(fig_dir, "fig_paul_realdata.png"), fig,
       width = 8.5, height = 3.4, dpi = 220, bg = "white")

fig
```

## Caveats

* **Cluster preservation is measured against the *raw* clustering**, not
  against published Paul 2015 cell-type annotations. The original Paul
  annotations (MetaCell 1-19 clusters) are not in the GEO supplementary
  files and were not retrieved here. ARI against raw measures whether
  imputation preserves the structure visible in the data, not whether
  it recovers ground truth.
* The TF-target list is small (<= 9 TFs, <= ~30 known pairs after
  intersecting with the variable-gene set) and human-curated. AUC is
  therefore noisy; a larger curated database (e.g. TRRUST,
  hematopoiesis-specific ChIP-seq targets) would give a tighter
  estimate.
* Cells were subsampled to **1 500** of the ~10 000 in the original
  paper, and we kept the **2 000** highest-variance genes. Increase
  `n_cells_keep` / `n_genes_keep` for the full benchmark; runtime
  scales roughly linearly in cells and in genes.
* **FACS-protein correlation**: the third metric named in the
  article, does not apply: Paul 2015 has no paired protein
  measurements per cell. The natural dataset for that comparison is
  the original MAGIC paper's HMLE TGF-beta EMT data with CD44 FACS
  (GSE114397).
