---
title: "Per-gene adaptive t: does removing the clustering step help?"
output: html_notebook
---

The article's per-module variant (`magic_modules`) groups genes via k-means
on PC-projected expression profiles, then picks a single $t$ per cluster
from the *averaged* per-gene MCV losses inside that cluster. On every test
we have run it has either tied or slightly trailed global MCV.

This notebook tests a simpler alternative: **per-gene adaptive $t$**. We
still compute the per-gene MCV loss curve (via the same `mcv_per_gene`
internal as `magic_modules`), but we skip the clustering and the
within-cluster averaging entirely - each gene gets its own $t^\star_g =
\arg\min_t \mathcal{L}_g(t)$. Two MCV loss flavours are compared:

* **MSE** - the standard count-scale Batson loss.
* **Pearson** - squared standardised residuals $(B - \hat\mu)^2 /
  \max(\hat\mu, \varepsilon)$, scale-invariant per gene.

Tests:

1. Synthetic *mixed-scale* regime (the regime that motivated per-module
   in the first place: gene groups with mean-expression scales $1:5:20$).
2. Paul 2015 bone-marrow data (`paul_realdata.Rmd` setup).

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(cvMAGIC)
  library(Rmagic)
  library(Matrix)
  library(Seurat)
  library(mclust)
  library(pROC)
  library(ggplot2)
  library(dplyr)
  library(patchwork)
})

fig_dir  <- normalizePath(file.path("..", "figures"), mustWork = FALSE)
data_dir <- normalizePath(file.path("..", "data"),    mustWork = FALSE)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

stopifnot("magic_per_gene" %in% getNamespaceExports("cvMAGIC"))
```

## Part 1 - synthetic mixed-scale regime

```{r synthetic}
make_mixed_counts <- function(n_cells, n_genes, n_clusters, n_groups,
                              scales, seed, libsize = 1500) {
  set.seed(seed)
  stopifnot(n_genes %% n_groups == 0)
  per_group <- n_genes %/% n_groups
  group <- rep(seq_len(n_groups), each = per_group)
  labels <- sample.int(n_clusters, n_cells, replace = TRUE)
  cent <- matrix(0.1, nrow = n_clusters, ncol = n_genes)
  for (g in seq_len(n_groups)) {
    gcols <- which(group == g)
    for (c in seq_len(n_clusters)) {
      sg <- sample(gcols, per_group %/% n_clusters)
      cent[c, sg] <- runif(length(sg), 2, 5) * scales[g]
    }
  }
  lam <- cent[labels, , drop = FALSE]
  lam <- lam / rowSums(lam) * libsize
  counts <- matrix(rpois(n_cells * n_genes, lam), n_cells, n_genes)
  list(counts = counts, lambda = lam, labels = labels, group = group)
}

mse <- function(a, b) mean((a - b)^2)

NPCA <- 14L; K <- 22L
d <- make_mixed_counts(300L, 120L, 3L, 3L,
                       scales = c(0.3, 1.5, 6.0), seed = 101L)
```

```{r synthetic-methods, message=FALSE}
results_syn <- list()

# cv_mcv (global)
sel <- mcv_select_t(d$counts, t_values = 1:12,
                    npca = NPCA, k = K, seed = 1L)
g <- magic_graph(d$counts, npca = NPCA, k = K)
imp <- magic_impute(d$counts, t = sel$t, graph = g)
results_syn$cv_mcv <- list(imp = imp, t = sel$t)

# cv_per_module
out <- magic_modules(d$counts, n_modules = 6L, t_values = 1:12,
                     npca = NPCA, k = K, seed = 1L)
results_syn$cv_per_module <- list(imp = out$imputed,
                                  t = out$t_per_module)

# cv_per_gene (MSE)
out <- magic_per_gene(d$counts, t_values = 1:12, loss = "mse",
                      npca = NPCA, k = K, seed = 1L)
results_syn$cv_per_gene_mse <- list(imp = out$imputed,
                                    t = out$t_per_gene)

# cv_per_gene (Pearson)
out <- magic_per_gene(d$counts, t_values = 1:12, loss = "pearson",
                      npca = NPCA, k = K, seed = 1L)
results_syn$cv_per_gene_pearson <- list(imp = out$imputed,
                                        t = out$t_per_gene)

# rmagic_auto
X <- d$counts; colnames(X) <- paste0("g", seq_len(ncol(X)))
out <- Rmagic::magic(X, t = "auto", knn = K, npca = NPCA, verbose = 0)
results_syn$rmagic_auto <- list(imp = as.matrix(out$result), t = NA)

syn_table <- data.frame(
  method = names(results_syn),
  mse    = sapply(results_syn, function(r) mse(r$imp, d$lambda)),
  median_t = sapply(results_syn, function(r)
    if (length(r$t) > 1) median(r$t, na.rm = TRUE) else r$t)
)
print(syn_table, row.names = FALSE)
```

The MSE here is what we want to minimise. Per-module ties global MCV;
per-gene (either flavour) is the test.

## Part 2 - Paul 2015 real-data

Re-runs the same comparison on the Paul bone-marrow subset used in
`paul_realdata.Rmd`. Data download is cached at `../data/`.

```{r paul-load, message=FALSE}
url      <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE72nnn/GSE72857/suppl/GSE72857_umitab.txt.gz"
gz_path  <- file.path(data_dir, "GSE72857_umitab.txt.gz")
txt_path <- file.path(data_dir, "GSE72857_umitab.txt")
if (!file.exists(gz_path))  {
  options(timeout = 600); utils::download.file(url, gz_path, mode = "wb",
                                               quiet = TRUE)
}
if (!file.exists(txt_path)) R.utils::gunzip(gz_path, destname = txt_path,
                                            remove = FALSE)

counts_full <- data.table::fread(txt_path, sep = "\t",
                                 header = TRUE, data.table = FALSE)
gene_names  <- counts_full[[1]]
counts_full <- as.matrix(counts_full[, -1, drop = FALSE])
rownames(counts_full) <- gene_names

libsize    <- colSums(counts_full)
keep_cell  <- libsize >= 500 & libsize <= quantile(libsize, 0.99)
set.seed(101L)
cell_idx   <- sample(which(keep_cell), 1500L)
cnts       <- counts_full[, cell_idx]
gene_pos   <- rowMeans(cnts > 0)
cnts       <- cnts[gene_pos >= 0.05, ]
gene_var   <- apply(cnts, 1, var)
gene_idx   <- order(gene_var, decreasing = TRUE)[seq_len(2000L)]
cnts       <- cnts[gene_idx, ]

X <- t(cnts)
storage.mode(X) <- "integer"
colnames(X) <- rownames(cnts)
rownames(X) <- colnames(cnts)
dim(X)
```

```{r paul-methods, message=FALSE}
NPCA_P <- 20L; K_P <- 30L
results_paul <- list()

# cv_mcv
sel <- mcv_select_t(X, t_values = 1:10,
                    npca = NPCA_P, k = K_P, seed = 1L)
g <- magic_graph(X, npca = NPCA_P, k = K_P)
results_paul$cv_mcv <- list(imp = magic_impute(X, t = sel$t, graph = g),
                            t = sel$t)

# cv_per_module
out <- magic_modules(X, n_modules = 8L, t_values = 1:10,
                     npca = NPCA_P, k = K_P, seed = 1L)
results_paul$cv_per_module <- list(imp = out$imputed,
                                   t = out$t_per_module)

# cv_per_gene (MSE)
out <- magic_per_gene(X, t_values = 1:10, loss = "mse",
                      npca = NPCA_P, k = K_P, seed = 1L)
results_paul$cv_per_gene_mse <- list(imp = out$imputed,
                                     t = out$t_per_gene)

# cv_per_gene (Pearson)
out <- magic_per_gene(X, t_values = 1:10, loss = "pearson",
                      npca = NPCA_P, k = K_P, seed = 1L)
results_paul$cv_per_gene_pearson <- list(imp = out$imputed,
                                         t = out$t_per_gene)

# rmagic_auto
out <- Rmagic::magic(X, t = "auto", knn = K_P, npca = NPCA_P, verbose = 0)
results_paul$rmagic_auto <- list(imp = as.matrix(out$result), t = NA)

# attach colnames
for (m in names(results_paul))
  colnames(results_paul[[m]]$imp) <- colnames(X)
```

```{r paul-t-distribution, fig.width = 7, fig.height = 2.6}
t_pg_mse     <- results_paul$cv_per_gene_mse$t
t_pg_pearson <- results_paul$cv_per_gene_pearson$t
t_df <- data.frame(
  method = factor(rep(c("per-gene MSE", "per-gene Pearson"),
                      times = c(length(t_pg_mse), length(t_pg_pearson))),
                  levels = c("per-gene MSE", "per-gene Pearson")),
  t      = c(t_pg_mse, t_pg_pearson)
)

ggplot(t_df, aes(t, fill = method)) +
  geom_bar(width = 0.7, colour = "black", linewidth = 0.2) +
  facet_wrap(~ method) +
  scale_fill_manual(values = c("per-gene MSE" = "#1f77b4",
                               "per-gene Pearson" = "#2ca02c"),
                    guide = "none") +
  labs(x = "Per-gene optimal t", y = "number of genes",
       title = "Distribution of per-gene t* on Paul 2015") +
  theme_classic(base_size = 9)
```

```{r paul-cluster}
cluster_louvain <- function(mat, npca = 20L, k_param = 20L, res = 0.6) {
  rownames(mat) <- paste0("cell", seq_len(nrow(mat)))
  colnames(mat) <- paste0("g",   seq_len(ncol(mat)))
  obj <- CreateSeuratObject(counts = t(mat), assay = "RNA")
  obj <- SetAssayData(obj, assay = "RNA", layer = "data",
                      new.data = log1p(t(mat)))
  VariableFeatures(obj) <- colnames(mat)
  obj <- ScaleData(obj, verbose = FALSE)
  obj <- RunPCA(obj, npcs = npca, verbose = FALSE,
                features = VariableFeatures(obj))
  obj <- FindNeighbors(obj, dims = seq_len(npca), k.param = k_param,
                       verbose = FALSE)
  obj <- FindClusters(obj, resolution = res, verbose = FALSE,
                      random.seed = 1L)
  as.integer(as.character(Idents(obj)))
}
cl_raw <- cluster_louvain(X)
cl_each <- lapply(results_paul, function(r) cluster_louvain(r$imp))
ari <- sapply(cl_each, adjustedRandIndex, cl_raw)
```

```{r paul-tf-auc}
tf_targets <- list(
  Gata1 = c("Klf1","Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1",
            "Slc4a1","Gypa","Epor"),
  Gata2 = c("Gata1","Klf1","Itga2b"),
  Klf1  = c("Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1","Slc4a1"),
  Sfpi1 = c("Cebpa","Mpo","Csf1r","Csf2ra","Csf2rb"),
  Cebpa = c("Mpo","Elane","Prtn3","Ctsg","Csf3r"),
  Cebpe = c("Mpo","Elane","Prtn3","Ctsg"),
  Gfi1  = c("Csf1r","Elane"),
  Irf8  = c("Cxcr4"),
  Tal1  = c("Gata1","Klf1","Gata2")
)
gene_set <- colnames(X)
score_method <- function(imp_mat) {
  pos <- list(); neg <- list()
  set.seed(7L)
  for (tf in names(tf_targets)) {
    if (!(tf %in% gene_set)) next
    targets <- intersect(tf_targets[[tf]], gene_set)
    if (length(targets) == 0L) next
    pool <- setdiff(gene_set, c(tf, targets))
    non_targets <- sample(pool, min(length(pool), 5L * length(targets)))
    tf_vec <- imp_mat[, tf]
    cor_pos <- abs(apply(imp_mat[, targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cor_neg <- abs(apply(imp_mat[, non_targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cor_pos[is.na(cor_pos)] <- 0
    cor_neg[is.na(cor_neg)] <- 0
    pos[[tf]] <- data.frame(score = cor_pos, label = 1L)
    neg[[tf]] <- data.frame(score = cor_neg, label = 0L)
  }
  pairs <- rbind(do.call(rbind, pos), do.call(rbind, neg))
  as.numeric(pROC::roc(pairs$label, pairs$score,
                       levels = c(0L, 1L), direction = "<",
                       quiet = TRUE)$auc)
}
aucs <- c(raw = score_method(X),
          sapply(results_paul, function(r) score_method(r$imp)))
```

```{r paul-summary}
paul_table <- data.frame(
  method   = names(results_paul),
  n_clust  = sapply(cl_each, function(c) length(unique(c))),
  ARI_raw  = ari,
  AUC      = aucs[names(results_paul)]
)
cat("Raw AUC = ", round(aucs["raw"], 3), "\n", sep = "")
print(paul_table, row.names = FALSE)
```

## Combined comparison figure

```{r combined-figure, fig.width = 10, fig.height = 6.8}
syn_plot <- syn_table |>
  mutate(method = factor(method,
                         levels = c("rmagic_auto", "cv_mcv",
                                    "cv_per_module",
                                    "cv_per_gene_mse",
                                    "cv_per_gene_pearson"))) |>
  arrange(method)

palette_all <- c(
  "rmagic_auto"        = "#ff7f0e",
  "cv_mcv"             = "#1f77b4",
  "cv_per_module"      = "#5fa8d3",
  "cv_per_gene_mse"    = "#2ca02c",
  "cv_per_gene_pearson"= "#8ed18a"
)

p_syn <- ggplot(syn_plot, aes(method, mse, fill = method)) +
  geom_col(width = 0.6, colour = "black", linewidth = 0.25) +
  geom_text(aes(label = sprintf("%.3f", mse)),
            vjust = -0.4, size = 3) +
  scale_fill_manual(values = palette_all, guide = "none") +
  scale_y_continuous(limits = c(0, max(syn_plot$mse) * 1.18),
                     expand = c(0, 0)) +
  labs(x = NULL, y = expression(MSE ~ "vs." ~ lambda),
       title = "(a) Synthetic mixed-scale: MSE vs known truth") +
  theme_classic(base_size = 10) +
  theme(plot.title = element_text(size = 11),
        axis.text.x = element_text(angle = 25, hjust = 1))

paul_plot <- paul_table |>
  mutate(method = factor(method, levels = levels(syn_plot$method))) |>
  arrange(method)

p_ari <- ggplot(paul_plot, aes(method, ARI_raw, fill = method)) +
  geom_col(width = 0.6, colour = "black", linewidth = 0.25) +
  geom_text(aes(label = sprintf("%.3f", ARI_raw)),
            vjust = -0.4, size = 3) +
  scale_fill_manual(values = palette_all, guide = "none") +
  scale_y_continuous(limits = c(0, max(paul_plot$ARI_raw) * 1.18),
                     expand = c(0, 0)) +
  labs(x = NULL, y = "ARI (vs raw clustering)",
       title = "(b) Paul 2015: cluster preservation") +
  theme_classic(base_size = 10) +
  theme(plot.title = element_text(size = 11),
        axis.text.x = element_text(angle = 25, hjust = 1))

p_auc <- ggplot(paul_plot, aes(method, AUC, fill = method)) +
  geom_col(width = 0.6, colour = "black", linewidth = 0.25) +
  geom_hline(yintercept = aucs["raw"], linetype = "dashed",
             colour = "grey50") +
  annotate("text", x = 0.6, y = aucs["raw"] + 0.005,
           label = sprintf("raw = %.3f", aucs["raw"]),
           hjust = 0, size = 3, colour = "grey30") +
  geom_text(aes(label = sprintf("%.3f", AUC)),
            vjust = -0.4, size = 3) +
  scale_fill_manual(values = palette_all, guide = "none") +
  coord_cartesian(ylim = c(min(paul_plot$AUC, aucs["raw"]) - 0.02,
                           max(paul_plot$AUC) + 0.015)) +
  labs(x = NULL, y = "ROC AUC",
       title = "(c) Paul 2015: TF-target prediction") +
  theme_classic(base_size = 10) +
  theme(plot.title = element_text(size = 11),
        axis.text.x = element_text(angle = 25, hjust = 1))

fig <- (p_syn | p_ari | p_auc) + plot_layout(widths = c(1, 1, 1))

ggsave(file.path(fig_dir, "fig_per_gene_comparison.png"), fig,
       width = 10, height = 3.6, dpi = 220, bg = "white")

fig
```
