---
title: "Tabula Muris marrow (10x droplet), third-dataset check of tolerant MCV"
output: html_notebook
---

`magic_tolerant` (tolerance-regularised MCV, eps = 5%) was the one selector
that beat upstream MAGIC on *both* benchmark datasets so far:

| Method                     | Paul AUC | PBMC AUC | Mean |
|----------------------------|----------|----------|------|
| rmagic_auto (upstream)     | 0.971    | 0.846    | 0.909 |
| cv_mcv vanilla (0:10)      | 0.973    | 0.856    | 0.914 |
| **magic_tolerant (eps=5%)**  | 0.977    | 0.856    | 0.917 |

Data: figshare droplet release (Tabula Muris v2), channel
`Marrow-10X_P7_2`, extracted to
`data/droplet/Marrow-10X_P7_2/{matrix.mtx,genes.tsv,barcodes.tsv}`.

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(cvMAGIC); library(Rmagic); library(Matrix); library(mclust)
  library(pROC); library(ggplot2); library(dplyr); library(tidyr)
  library(patchwork)
})
stopifnot(Rmagic::pymagic_is_available())
stopifnot("magic_tolerant" %in% getNamespaceExports("cvMAGIC"))

fig_dir  <- normalizePath(file.path("..", "figures"), mustWork = FALSE)
data_dir <- normalizePath(file.path("..", "data"),    mustWork = FALSE)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

## Load the 10x droplet marrow channel

`genes.tsv` here carries the gene symbol in *both* columns (MGI symbols,
identical to Paul's gene names), so no Ensembl-symbol mapping is needed.

```{r load, message=FALSE}
mtx_dir <- file.path(data_dir, "droplet", "Marrow-10X_P7_2")
mat <- Matrix::readMM(file.path(mtx_dir, "matrix.mtx"))           # genes x cells
genes <- read.table(file.path(mtx_dir, "genes.tsv"), sep = "\t",
                    header = FALSE, stringsAsFactors = FALSE)
rownames(mat) <- make.unique(genes$V1)
colnames(mat) <- readLines(file.path(mtx_dir, "barcodes.tsv"))
cat("Raw matrix:", nrow(mat), "genes x", ncol(mat), "cells\n")
cat("Median UMIs/cell:", median(Matrix::colSums(mat)),
    " | median genes/cell:", median(Matrix::colSums(mat > 0)), "\n")
```

## Filter + subsample (identical protocol to Paul / PBMC notebooks)

```{r filter, message=FALSE}
libsize   <- Matrix::colSums(mat)
keep_cell <- libsize >= 500 & libsize <= quantile(libsize, 0.99)
cat("Cells passing libsize filter:", sum(keep_cell), "/",
    length(keep_cell), "\n")

set.seed(101L)
cell_idx <- sample(which(keep_cell), min(1500L, sum(keep_cell)))
cnts <- mat[, cell_idx, drop = FALSE]

# Mouse hematopoietic TF-target programme (same list as the Paul notebook).
tf_targets <- list(
  Gata1 = c("Klf1","Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1",
            "Slc4a1","Gypa","Epor"),
  Gata2 = c("Gata1","Klf1","Itga2b"),
  Klf1  = c("Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1","Slc4a1"),
  Sfpi1 = c("Cebpa","Mpo","Csf1r","Csf2ra","Csf2rb"),
  Cebpa = c("Mpo","Elane","Prtn3","Ctsg","Csf3r"),
  Cebpe = c("Mpo","Elane","Prtn3","Ctsg"),
  Gfi1  = c("Csf1r","Elane"),
  Irf8  = c("Cxcr4"),
  Tal1  = c("Gata1","Klf1","Gata2"))
must_keep <- intersect(unique(c(names(tf_targets), unlist(tf_targets))),
                       rownames(cnts))

# 10x-appropriate expression filter (>=3% of cells), then top-2000 HVG
# unioned with the force-kept TF/target genes.
gene_pos <- Matrix::rowMeans(cnts > 0)
cnts <- cnts[gene_pos >= 0.03 | rownames(cnts) %in% must_keep, , drop = FALSE]
gene_var <- apply(cnts, 1, function(x) var(as.numeric(x)))
hv_idx <- order(gene_var, decreasing = TRUE)[seq_len(min(2000L, length(gene_var)))]
keep_genes <- union(rownames(cnts)[hv_idx], must_keep)
cnts <- cnts[keep_genes, , drop = FALSE]

X <- as.matrix(t(cnts)); storage.mode(X) <- "integer"
colnames(X) <- rownames(cnts); rownames(X) <- colnames(cnts)
cat("Final matrix:", nrow(X), "cells x", ncol(X), "genes\n")
present_tfs <- intersect(names(tf_targets), colnames(X))
cat("TFs present:", length(present_tfs), "/", length(tf_targets),
    "(", paste(present_tfs, collapse = ","), ")\n")
rm(mat); invisible(gc())
```

## Scoring: TF-target Spearman-correlation ROC AUC

Same scorer as the Paul and PBMC benchmarks: for each TF, positive pairs
are its curated targets and negatives are 5x as many random non-target
genes; the score is |Spearman| between imputed TF and gene; AUC is over
the pooled pos/neg pairs.

```{r score-fn}
score_method <- function(imp_mat, tf_targets, gene_set) {
  pos <- list(); neg <- list()
  set.seed(7L)
  for (tf in names(tf_targets)) {
    if (!(tf %in% gene_set)) next
    targets <- intersect(tf_targets[[tf]], gene_set)
    if (length(targets) == 0L) next
    pool <- setdiff(gene_set, c(tf, targets))
    non_targets <- sample(pool, min(length(pool), 5L * length(targets)))
    tf_vec <- imp_mat[, tf]
    cor_pos <- abs(apply(imp_mat[, targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cor_neg <- abs(apply(imp_mat[, non_targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cor_pos[is.na(cor_pos)] <- 0; cor_neg[is.na(cor_neg)] <- 0
    pos[[tf]] <- data.frame(score = cor_pos, label = 1L)
    neg[[tf]] <- data.frame(score = cor_neg, label = 0L)
  }
  pairs <- rbind(do.call(rbind, pos), do.call(rbind, neg))
  list(auc = as.numeric(pROC::roc(pairs$label, pairs$score,
                                  levels = c(0L, 1L), direction = "<",
                                  quiet = TRUE)$auc),
       n_pos = sum(pairs$label == 1L), n_neg = sum(pairs$label == 0L))
}

impute_at <- function(X, t, npca = 20L, k = 30L) {
  if (t == 0L) { out <- X } else {
    g <- magic_graph(X, npca = npca, k = k)
    out <- magic_impute(X, t = t, graph = g)
  }
  colnames(out) <- colnames(X); out
}
```

## Oracle t-sweep: what does the AUC-vs-t curve look like here?

This tells us the *regime*: a curve that rises with t (like Paul) means
sparse data wants smoothing; a curve that peaks at t=0 (like PBMC) means
dense data wants none. The tolerance rule's job is to track whichever it
is.

```{r sweep, message=FALSE}
t_grid <- 0:10
g_full <- magic_graph(X, npca = 20L, k = 30L)
auc_by_t <- sapply(t_grid, function(t) {
  imp <- if (t == 0L) X else magic_impute(X, t = t, graph = g_full)
  colnames(imp) <- colnames(X)
  score_method(imp, tf_targets, colnames(X))$auc
})
sweep_tbl <- data.frame(t = t_grid, AUC = auc_by_t)
oracle_t <- t_grid[which.max(auc_by_t)]
cat("AUC by global t:\n"); print(sweep_tbl, row.names = FALSE)
cat(sprintf("Oracle (best global t) = %d, AUC = %.3f\n",
            oracle_t, max(auc_by_t)))
```

## Head-to-head: upstream vs vanilla MCV vs tolerant MCV

```{r methods, message=FALSE}
auc_raw <- score_method(X, tf_targets, colnames(X))

# upstream MAGIC, t = "auto"
imp_rmagic <- as.matrix(Rmagic::magic(X, t = "auto", knn = 30L, npca = 20L,
                                      verbose = 0)$result)
colnames(imp_rmagic) <- colnames(X)
auc_rmagic <- score_method(imp_rmagic, tf_targets, colnames(X))$auc

# vanilla cv_mcv (argmin of MCV loss, search 0:10)
van <- mcv_select_t(X, t_values = 0:10, npca = 20L, k = 30L, seed = 1L)
auc_van <- score_method(impute_at(X, van$t), tf_targets, colnames(X))$auc

# shipped tolerant MCV (eps = 5%)
tol <- magic_tolerant(X, t_values = 0:10, tolerance = 0.05,
                      npca = 20L, k = 30L, seed = 1L)
imp_tol <- tol$imputed; colnames(imp_tol) <- colnames(X)
auc_tol <- score_method(imp_tol, tf_targets, colnames(X))$auc

summary_df <- data.frame(
  method = c("raw", "rmagic_auto", "cv_mcv (vanilla)", "magic_tolerant"),
  t      = c(NA, NA, van$t, tol$t),
  AUC    = c(auc_raw$auc, auc_rmagic, auc_van, auc_tol))

cat(sprintf("\nTabula Muris marrow (10x) TF-target AUC  (%d pos / %d neg pairs)\n",
            auc_raw$n_pos, auc_raw$n_neg))
print(summary_df, row.names = FALSE)
cat(sprintf("\nMCV argmin t = %d ; tolerant t = %d ; oracle t = %d\n",
            van$t, tol$t, oracle_t))
cat(sprintf("tolerant vs upstream rmagic_auto: %+.3f AUC\n",
            auc_tol - auc_rmagic))
cat(sprintf("tolerant vs vanilla cv_mcv:       %+.3f AUC\n",
            auc_tol - auc_van))
```

## Figure

```{r figure, fig.width = 9, fig.height = 4}
p_sweep <- ggplot(sweep_tbl, aes(t, AUC)) +
  geom_line(linewidth = 0.8, colour = "grey40") +
  geom_point(size = 2, colour = "grey40") +
  geom_point(data = subset(sweep_tbl, t == van$t),
             aes(colour = "vanilla MCV argmin"), size = 4) +
  geom_point(data = subset(sweep_tbl, t == tol$t),
             aes(colour = "tolerant MCV (eps=5%)"), size = 3) +
  geom_point(data = subset(sweep_tbl, t == oracle_t),
             aes(colour = "oracle"), size = 2) +
  scale_colour_manual(values = c("vanilla MCV argmin" = "#1f77b4",
                                 "tolerant MCV (eps=5%)" = "#2ca02c",
                                 "oracle" = "#d62728"), name = NULL) +
  scale_x_continuous(breaks = t_grid) +
  labs(x = "global diffusion time t", y = "TF-target AUC",
       title = "(a) AUC-vs-t and where each selector lands") +
  theme_classic(base_size = 10) + theme(legend.position = "bottom")

bar_df <- summary_df
bar_df$method <- factor(bar_df$method,
  levels = c("raw", "rmagic_auto", "cv_mcv (vanilla)", "magic_tolerant"))
p_bar <- ggplot(bar_df, aes(method, AUC, fill = method)) +
  geom_col(width = 0.6, colour = "black", linewidth = 0.25) +
  geom_text(aes(label = ifelse(is.na(t), sprintf("%.3f", AUC),
                               sprintf("%.3f\n(t=%d)", AUC, t))),
            vjust = -0.15, size = 3, lineheight = 0.85) +
  scale_fill_manual(values = c("raw" = "grey70",
                               "rmagic_auto" = "#ff7f0e",
                               "cv_mcv (vanilla)" = "#1f77b4",
                               "magic_tolerant" = "#2ca02c"), guide = "none") +
  coord_cartesian(ylim = c(min(bar_df$AUC) - 0.02, max(bar_df$AUC) + 0.02)) +
  labs(x = NULL, y = "TF-target ROC AUC",
       title = "(b) Tabula Muris marrow: head-to-head") +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

fig <- p_sweep + p_bar + plot_layout(widths = c(1, 1))
ggsave(file.path(fig_dir, "fig_tabula_muris.png"), fig,
       width = 9, height = 4.2, dpi = 220, bg = "white")
fig
```

## Caveats

* Single channel (`Marrow-10X_P7_2`, ~1500 cells after subsampling).
  `Marrow-10X_P7_3` is extracted alongside and can be pooled or used as a
  replicate if a second run is wanted.
* Same scoring caveats as Paul/PBMC: the TF-target list is small and
  curated from lineage biology, not a ChIP-seq ground truth; absolute
  AUCs are comparative, not absolute.
* ARI-vs-raw clustering is omitted here (the TF-target AUC is the metric
  the tolerance rule was designed against); add it back if a cluster-
  preservation cross-check is wanted.
```
