---
title: "Stress-testing tolerance-regularised MCV: does it really enhance MAGIC?"
output: html_notebook
---

The tolerance rule (`mcv_select_t_tolerant`, eps=5%) only ever beat plain
MCV on **one** dataset (Paul 2015, +0.006); on the two dense datasets it
was inert (picked the same t=0). Its whole reason to exist is the
**sparse** regime. This notebook stress-tests exactly that claim two
ways:

* **Part A: controlled sparsity sweep.** Take Tabula Muris marrow
  (same mouse hematopoietic biology as Paul, same TF-target list) and
  binomially down-sample its UMIs across a depth ladder, holding the
  cells, the gene panel, and the biology *fixed*. The only thing that
  changes is sparsity. If the tolerance rule is principled, its benefit
  over plain MCV should *emerge as the data gets sparser* and *vanish
  when dense*. If the Paul win was a fluke, the benefit never appears.
* **Part B: Paul multi-seed robustness.** The entire tolerance claim
  rests on a single +0.006 from one random 1500-cell subsample. Re-run
  across many subsample seeds and ask whether the tolerant-vanilla and
  tolerant-rmagic gaps are reliably positive or just sampling noise.

Everything is measured against the same TF-target Spearman-AUC metric
and compared head-to-head with upstream `rmagic_auto`.

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(cvMAGIC); library(Rmagic); library(Matrix); library(pROC)
  library(ggplot2); library(dplyr); library(tidyr); library(patchwork)
})
stopifnot(Rmagic::pymagic_is_available())
fig_dir  <- normalizePath(file.path("..", "figures"), mustWork = FALSE)
data_dir <- normalizePath(file.path("..", "data"),    mustWork = FALSE)
out_dir  <- normalizePath(file.path("..", "notebooks"), mustWork = FALSE)

# Mouse hematopoietic TF->target programme (identical to Paul/TM notebooks)
tf_targets <- list(
  Gata1 = c("Klf1","Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1",
            "Slc4a1","Gypa","Epor"),
  Gata2 = c("Gata1","Klf1","Itga2b"),
  Klf1  = c("Hba-a1","Hba-a2","Hbb-b1","Hbb-b2","Aqp1","Slc4a1"),
  Sfpi1 = c("Cebpa","Mpo","Csf1r","Csf2ra","Csf2rb"),
  Cebpa = c("Mpo","Elane","Prtn3","Ctsg","Csf3r"),
  Cebpe = c("Mpo","Elane","Prtn3","Ctsg"),
  Gfi1  = c("Csf1r","Elane"),
  Irf8  = c("Cxcr4"),
  Tal1  = c("Gata1","Klf1","Gata2"))

score_method <- function(imp_mat, tf_targets, gene_set) {
  pos <- list(); neg <- list(); set.seed(7L)
  for (tf in names(tf_targets)) {
    if (!(tf %in% gene_set)) next
    targets <- intersect(tf_targets[[tf]], gene_set)
    if (length(targets) == 0L) next
    pool <- setdiff(gene_set, c(tf, targets))
    non_targets <- sample(pool, min(length(pool), 5L * length(targets)))
    tf_vec <- imp_mat[, tf]
    cp <- abs(apply(imp_mat[, targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cn <- abs(apply(imp_mat[, non_targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cp[is.na(cp)] <- 0; cn[is.na(cn)] <- 0
    pos[[tf]] <- data.frame(score = cp, label = 1L)
    neg[[tf]] <- data.frame(score = cn, label = 0L)
  }
  pairs <- rbind(do.call(rbind, pos), do.call(rbind, neg))
  as.numeric(pROC::roc(pairs$label, pairs$score, levels = c(0L, 1L),
                       direction = "<", quiet = TRUE)$auc)
}

# All three selectors + oracle for one count matrix, sharing one full graph.
eval_all <- function(X, tf_targets, t_values = 0:10, t_oracle = 0:15,
                     npca = 20L, k = 30L, seed = 1L) {
  gs <- colnames(X)
  g  <- magic_graph(X, npca = npca, k = k)
  auc_at <- function(t) {
    imp <- if (t == 0L) X else magic_impute(X, t = t, graph = g)
    colnames(imp) <- gs; score_method(imp, tf_targets, gs)
  }
  auc_curve <- sapply(t_oracle, auc_at); names(auc_curve) <- t_oracle
  sel <- mcv_select_t(X, t_values = t_values, npca = npca, k = k, seed = seed)
  thr <- min(sel$loss) * 1.05
  t_tol <- sel$t_values[max(which(sel$loss <= thr))]
  rmg <- as.matrix(Rmagic::magic(X, t = "auto", knn = k, npca = npca,
                                 verbose = 0)$result)
  colnames(rmg) <- gs
  data.frame(
    t_vanilla = sel$t,            auc_vanilla = auc_curve[as.character(sel$t)],
    t_tol     = t_tol,            auc_tol     = auc_curve[as.character(t_tol)],
    t_oracle  = as.integer(names(which.max(auc_curve))),
    auc_oracle = max(auc_curve),
    auc_rmagic = score_method(rmg, tf_targets, gs),
    auc_raw    = auc_curve["0"],
    row.names = NULL)
}

downsample <- function(X, p, seed) {
  set.seed(seed)
  Xd <- matrix(rbinom(length(X), as.vector(X), p),
               nrow = nrow(X), ncol = ncol(X),
               dimnames = dimnames(X))
  storage.mode(Xd) <- "integer"; Xd
}
```

## Part A: controlled sparsity sweep on Tabula Muris marrow

Build the fixed 1500-cell / 2000-HVG marrow panel once at full depth,
then down-sample only the counts.

```{r build-tm, message=FALSE}
mtx_dir <- file.path(data_dir, "droplet", "Marrow-10X_P7_2")
mat <- Matrix::readMM(file.path(mtx_dir, "matrix.mtx"))
g_tsv <- read.table(file.path(mtx_dir, "genes.tsv"), sep = "\t",
                    header = FALSE, stringsAsFactors = FALSE)
rownames(mat) <- make.unique(g_tsv$V1)
colnames(mat) <- readLines(file.path(mtx_dir, "barcodes.tsv"))
libsize   <- Matrix::colSums(mat)
keep_cell <- libsize >= 500 & libsize <= quantile(libsize, 0.99)
set.seed(101L)
cell_idx  <- sample(which(keep_cell), 1500L)
cnts <- mat[, cell_idx, drop = FALSE]
must_keep <- intersect(unique(c(names(tf_targets), unlist(tf_targets))),
                       rownames(cnts))
gene_pos  <- Matrix::rowMeans(cnts > 0)
cnts <- cnts[gene_pos >= 0.03 | rownames(cnts) %in% must_keep, , drop = FALSE]
gene_var  <- apply(cnts, 1, function(x) var(as.numeric(x)))
hv <- order(gene_var, decreasing = TRUE)[seq_len(min(2000L, length(gene_var)))]
keep_genes <- union(rownames(cnts)[hv], must_keep)
cnts <- cnts[keep_genes, , drop = FALSE]
X_full <- as.matrix(t(cnts)); storage.mode(X_full) <- "integer"
colnames(X_full) <- rownames(cnts); rownames(X_full) <- colnames(cnts)
rm(mat); invisible(gc())
cat("Full-depth panel:", nrow(X_full), "cells x", ncol(X_full),
    "genes; median UMI/cell =", median(rowSums(X_full)), "\n")
```

```{r sweep, message=FALSE}
depths <- c(0.05, 0.10, 0.25, 0.50, 1.00)
seeds  <- 1:3
sweep <- list()
for (p in depths) for (s in seeds) {
  Xd <- if (p == 1.0) X_full else downsample(X_full, p, seed = s)
  r  <- eval_all(Xd, tf_targets, seed = s)
  r$depth <- p; r$seed <- s; r$median_umi <- median(rowSums(Xd))
  sweep[[paste(p, s)]] <- r
  cat(sprintf("p=%.2f seed=%d  med_umi=%.0f | vanilla t=%d(%.3f) tol t=%d(%.3f) oracle t=%d(%.3f) rmagic=%.3f\n",
              p, s, r$median_umi, r$t_vanilla, r$auc_vanilla, r$t_tol,
              r$auc_tol, r$t_oracle, r$auc_oracle, r$auc_rmagic))
}
sweep_df <- do.call(rbind, sweep)
write.csv(sweep_df, file.path(out_dir, "tolerance_stress_sweep.csv"),
          row.names = FALSE)
```

```{r sweep-agg}
agg <- sweep_df |>
  group_by(median_umi = round(median_umi)) |>
  summarise(across(c(auc_vanilla, auc_tol, auc_oracle, auc_rmagic,
                     t_vanilla, t_tol, t_oracle),
                   list(m = mean, sd = sd)),
            gap_tol_van  = mean(auc_tol - auc_vanilla),
            gap_tol_van_sd = sd(auc_tol - auc_vanilla),
            gap_tol_rmg  = mean(auc_tol - auc_rmagic),
            .groups = "drop") |>
  arrange(median_umi)
cat("Sparsity sweep (mean over seeds), sparse -> dense:\n")
print(as.data.frame(agg[, c("median_umi","t_vanilla_m","t_tol_m","t_oracle_m",
                            "auc_vanilla_m","auc_tol_m","auc_oracle_m",
                            "auc_rmagic_m","gap_tol_van","gap_tol_van_sd")]),
      row.names = FALSE, digits = 3)
```

```{r sweep-fig, fig.width = 10, fig.height = 4}
long <- sweep_df |>
  pivot_longer(c(auc_vanilla, auc_tol, auc_oracle, auc_rmagic),
               names_to = "method", values_to = "AUC") |>
  mutate(method = recode(method,
           auc_vanilla = "cv_mcv (vanilla)", auc_tol = "magic_tolerant",
           auc_oracle = "oracle", auc_rmagic = "rmagic_auto"))
sumv <- long |> group_by(median_umi = round(median_umi), method) |>
  summarise(m = mean(AUC), sd = sd(AUC), .groups = "drop")

p1 <- ggplot(sumv, aes(median_umi, m, colour = method)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  geom_errorbar(aes(ymin = m - sd, ymax = m + sd), width = 0.03, alpha = .5) +
  scale_x_log10() +
  scale_colour_manual(values = c("cv_mcv (vanilla)" = "#1f77b4",
    "magic_tolerant" = "#2ca02c", "oracle" = "#d62728",
    "rmagic_auto" = "#ff7f0e"), name = NULL) +
  labs(x = "median UMI/cell (log; sparse -> dense)", y = "TF-target AUC",
       title = "(a) AUC vs sequencing depth") +
  theme_classic(base_size = 10) + theme(legend.position = "bottom")

gap <- sweep_df |> group_by(median_umi = round(median_umi)) |>
  summarise(gap = mean(auc_tol - auc_vanilla),
            sd = sd(auc_tol - auc_vanilla), .groups = "drop")
p2 <- ggplot(gap, aes(median_umi, gap)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = gap - sd, ymax = gap + sd), alpha = .2,
              fill = "#2ca02c") +
  geom_line(linewidth = 0.8, colour = "#2ca02c") +
  geom_point(size = 2, colour = "#2ca02c") +
  scale_x_log10() +
  labs(x = "median UMI/cell (log; sparse -> dense)",
       y = "AUC(tolerant) - AUC(vanilla)",
       title = "(b) Does tolerance help as data gets sparser?") +
  theme_classic(base_size = 10)

fig <- p1 + p2 + plot_layout(widths = c(1.1, 1))
ggsave(file.path(fig_dir, "fig_tolerance_stress_sparsity.png"), fig,
       width = 10, height = 4.2, dpi = 220, bg = "white")
fig
```

## Part B: Paul 2015 multi-seed robustness

Re-run the original sparse win across many cell-subsample seeds.

```{r paul, message=FALSE}
txt_path <- file.path(data_dir, "GSE72857_umitab.txt")
cf <- data.table::fread(txt_path, sep = "\t", header = TRUE, data.table = FALSE)
gn <- cf[[1]]; cf <- as.matrix(cf[, -1, drop = FALSE]); rownames(cf) <- gn
libsize <- colSums(cf)
keep_cell_paul <- which(libsize >= 500 & libsize <= quantile(libsize, 0.99))

paul_seeds <- 101:108
paul <- list()
for (s in paul_seeds) {
  set.seed(s)
  ci <- sample(keep_cell_paul, 1500L)
  cn <- cf[, ci]
  gp <- rowMeans(cn > 0); cn <- cn[gp >= 0.05, ]
  gv <- apply(cn, 1, var)
  gi <- order(gv, decreasing = TRUE)[seq_len(2000L)]
  cn <- cn[gi, ]
  Xp <- t(cn); storage.mode(Xp) <- "integer"
  colnames(Xp) <- rownames(cn); rownames(Xp) <- colnames(cn)
  r <- eval_all(Xp, tf_targets, seed = 1L)
  r$seed <- s; paul[[as.character(s)]] <- r
  cat(sprintf("Paul seed=%d | vanilla t=%d(%.3f) tol t=%d(%.3f) oracle t=%d(%.3f) rmagic=%.3f\n",
              s, r$t_vanilla, r$auc_vanilla, r$t_tol, r$auc_tol,
              r$t_oracle, r$auc_oracle, r$auc_rmagic))
}
paul_df <- do.call(rbind, paul)
write.csv(paul_df, file.path(out_dir, "tolerance_stress_paul.csv"),
          row.names = FALSE)
rm(cf); invisible(gc())
```

```{r paul-stats}
g_tv <- paul_df$auc_tol - paul_df$auc_vanilla   # tolerant vs vanilla MCV
g_tr <- paul_df$auc_tol - paul_df$auc_rmagic    # tolerant vs upstream MAGIC
report <- function(x, lbl) {
  tt <- tryCatch(t.test(x), error = function(e) NULL)
  cat(sprintf("%-28s mean=%+.4f sd=%.4f  >0 in %d/%d seeds%s\n",
      lbl, mean(x), sd(x), sum(x > 0), length(x),
      if (!is.null(tt)) sprintf("  (paired t p=%.3f, 95%% CI [%+.4f, %+.4f])",
        tt$p.value, tt$conf.int[1], tt$conf.int[2]) else ""))
}
cat("Paul 2015 robustness across", length(paul_seeds), "subsample seeds:\n")
report(g_tv, "tolerant - vanilla MCV")
report(g_tr, "tolerant - rmagic_auto")
cat(sprintf("\nMedian chosen t: vanilla=%g, tolerant=%g, oracle=%g\n",
            median(paul_df$t_vanilla), median(paul_df$t_tol),
            median(paul_df$t_oracle)))
```

```{r paul-fig, fig.width = 7, fig.height = 4}
pl <- paul_df |>
  select(seed, auc_rmagic, auc_vanilla, auc_tol, auc_oracle) |>
  pivot_longer(-seed, names_to = "method", values_to = "AUC") |>
  mutate(method = recode(method, auc_rmagic = "rmagic_auto",
           auc_vanilla = "cv_mcv (vanilla)", auc_tol = "magic_tolerant",
           auc_oracle = "oracle"),
         method = factor(method, levels = c("rmagic_auto",
           "cv_mcv (vanilla)", "magic_tolerant", "oracle")))
fig2 <- ggplot(pl, aes(method, AUC, fill = method)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = .5) +
  geom_jitter(width = 0.12, size = 1.6, alpha = .8) +
  scale_fill_manual(values = c("rmagic_auto" = "#ff7f0e",
    "cv_mcv (vanilla)" = "#1f77b4", "magic_tolerant" = "#2ca02c",
    "oracle" = "#d62728"), guide = "none") +
  labs(x = NULL, y = "TF-target AUC",
       title = sprintf("Paul 2015: %d-seed robustness", length(paul_seeds))) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
ggsave(file.path(fig_dir, "fig_tolerance_stress_paul.png"), fig2,
       width = 7, height = 4, dpi = 220, bg = "white")
fig2
```

## Read-out

* **Part A** confirms (or refutes) the *mechanism*: the tolerant-vanilla
  gap in panel (b) should be ~0 at high depth and grow positive as
  median UMI/cell falls, with the chosen t tracking the oracle upward.
* **Part B** confirms (or refutes) the *headline*: whether the Paul win
  is reproducible across subsamples (CI excludes 0) or sampling noise.
