---
title: "Tolerance rule on a sparse UMI trajectory, Tusi 2018 inDrop HSPCs"
output: html_notebook
---

The stress test showed the tolerance rule's benefit needs a regime that
is **sparse AND a differentiation trajectory** (the only place the AUC
keeps rising past MCV's argmin). Paul 2015 is that regime; the dense 10x
sets are not, and down-sampling a *discrete* tissue (TM marrow) doesn't
recreate it (oracle stays t=0). This notebook gives the rule its fairest
possible shot on a **second, independent** dataset that is both:

* **UMI** (inDrop), so MCV's Poisson-thinning assumption is valid
  (unlike Smart-seq2 read counts), and
* a **continuous HSPC differentiation trajectory**: mouse Kit+ bone
  marrow progenitors branching into erythroid / myeloid / basophil fates
  (Tusi et al., *Nature* 2018; GSE89754, `basal_bone_marrow`).

Same mouse hematopoietic TF-target metric as Paul (symbols adjusted to
this dataset: `Sfpi1`-`Spi1`, `Hbb-b1/2`-`Hbb-bs/bt`). The question:

1. Is the oracle t > 0 here (does smoothing actually help)?
2. If so, does the tolerance rule track toward it and beat both vanilla
   MCV and upstream `rmagic_auto`: reproducibly across seeds?

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(cvMAGIC); library(Rmagic); library(Matrix); library(pROC)
  library(ggplot2); library(dplyr); library(tidyr); library(patchwork)
})
stopifnot(Rmagic::pymagic_is_available())
fig_dir  <- normalizePath(file.path("..", "figures"), mustWork = FALSE)
data_dir <- normalizePath(file.path("..", "data"),    mustWork = FALSE)
out_dir  <- normalizePath(file.path("..", "notebooks"), mustWork = FALSE)

tf_targets <- list(
  Gata1 = c("Klf1","Hba-a1","Hba-a2","Hbb-bs","Hbb-bt","Aqp1",
            "Slc4a1","Gypa","Epor"),
  Gata2 = c("Gata1","Klf1","Itga2b"),
  Klf1  = c("Hba-a1","Hba-a2","Hbb-bs","Hbb-bt","Aqp1","Slc4a1"),
  Spi1  = c("Cebpa","Mpo","Csf1r","Csf2ra","Csf2rb"),
  Cebpa = c("Mpo","Elane","Prtn3","Ctsg","Csf3r"),
  Cebpe = c("Mpo","Elane","Prtn3","Ctsg"),
  Gfi1  = c("Csf1r","Elane"),
  Irf8  = c("Cxcr4"),
  Tal1  = c("Gata1","Klf1","Gata2"))

score_method <- function(imp_mat, tf_targets, gene_set) {
  pos <- list(); neg <- list(); set.seed(7L)
  for (tf in names(tf_targets)) {
    if (!(tf %in% gene_set)) next
    targets <- intersect(tf_targets[[tf]], gene_set)
    if (length(targets) == 0L) next
    pool <- setdiff(gene_set, c(tf, targets))
    non_targets <- sample(pool, min(length(pool), 5L * length(targets)))
    tf_vec <- imp_mat[, tf]
    cp <- abs(apply(imp_mat[, targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cn <- abs(apply(imp_mat[, non_targets, drop = FALSE], 2, function(g)
      suppressWarnings(cor(tf_vec, g, method = "spearman"))))
    cp[is.na(cp)] <- 0; cn[is.na(cn)] <- 0
    pos[[tf]] <- data.frame(score = cp, label = 1L)
    neg[[tf]] <- data.frame(score = cn, label = 0L)
  }
  pairs <- rbind(do.call(rbind, pos), do.call(rbind, neg))
  list(auc = as.numeric(pROC::roc(pairs$label, pairs$score,
         levels = c(0L, 1L), direction = "<", quiet = TRUE)$auc),
       n_pos = sum(pairs$label == 1L), n_neg = sum(pairs$label == 0L))
}

eval_all <- function(X, tf_targets, t_values = 0:10, t_oracle = 0:15,
                     npca = 20L, k = 30L, seed = 1L) {
  gs <- colnames(X); g <- magic_graph(X, npca = npca, k = k)
  auc_at <- function(t) {
    imp <- if (t == 0L) X else magic_impute(X, t = t, graph = g)
    colnames(imp) <- gs; score_method(imp, tf_targets, gs)$auc
  }
  curve <- sapply(t_oracle, auc_at); names(curve) <- t_oracle
  sel <- mcv_select_t(X, t_values = t_values, npca = npca, k = k, seed = seed)
  t_tol <- sel$t_values[max(which(sel$loss <= min(sel$loss) * 1.05))]
  rmg <- as.matrix(Rmagic::magic(X, t = "auto", knn = k, npca = npca,
                                 verbose = 0)$result); colnames(rmg) <- gs
  list(curve = curve,
       row = data.frame(
         t_vanilla = sel$t,  auc_vanilla = curve[as.character(sel$t)],
         t_tol = t_tol,      auc_tol     = curve[as.character(t_tol)],
         t_oracle = as.integer(names(which.max(curve))),
         auc_oracle = max(curve),
         auc_rmagic = score_method(rmg, tf_targets, gs)$auc,
         auc_raw = curve["0"], row.names = NULL))
}
```

## Load the basal bone-marrow inDrop UMI matrix

CSV is cells x genes with five leading metadata columns
(`cell_id, barcode, library_id, seq_run_id, pass_filter`).

```{r load, message=FALSE}
csv <- file.path(data_dir, "GSM2388072_basal_bone_marrow.raw_umifm_counts.csv")
df <- data.table::fread(csv, sep = ",", header = TRUE, data.table = FALSE)
meta_cols <- c("cell_id","barcode","library_id","seq_run_id","pass_filter")
cell_id <- df$cell_id
M <- as.matrix(df[, setdiff(colnames(df), meta_cols), drop = FALSE])
rownames(M) <- cell_id
storage.mode(M) <- "integer"            # raw UMI counts, cells x genes
rm(df); invisible(gc())
cat("Raw inDrop matrix:", nrow(M), "cells x", ncol(M), "genes\n")
cat("Median UMIs/cell:", median(rowSums(M)),
    "| median genes/cell:", median(rowSums(M > 0)), "\n")
```

## Build the panel (same protocol as Paul / TM: 1500 cells, 2000 HVG)

```{r panel, message=FALSE}
build_panel <- function(M, tf_targets, seed = 101L, n_cells = 1500L,
                        n_genes = 2000L) {
  libsize <- rowSums(M)
  keep <- which(libsize >= 1000 & libsize <= quantile(libsize, 0.99))
  set.seed(seed)
  ci <- sample(keep, min(n_cells, length(keep)))
  cn <- M[ci, , drop = FALSE]
  must_keep <- intersect(unique(c(names(tf_targets), unlist(tf_targets))),
                         colnames(cn))
  gpos <- colMeans(cn > 0)
  cn <- cn[, gpos >= 0.05 | colnames(cn) %in% must_keep, drop = FALSE]
  gv <- apply(cn, 2, var)
  hv <- order(gv, decreasing = TRUE)[seq_len(min(n_genes, length(gv)))]
  keep_genes <- union(colnames(cn)[hv], must_keep)
  X <- cn[, keep_genes, drop = FALSE]; storage.mode(X) <- "integer"; X
}
X0 <- build_panel(M, tf_targets, seed = 101L)
cat("Panel:", nrow(X0), "cells x", ncol(X0), "genes; median UMI/cell =",
    median(rowSums(X0)), "\n")
present <- intersect(names(tf_targets), colnames(X0))
cat("TFs present:", length(present), "/", length(tf_targets),
    "(", paste(present, collapse = ","), ")\n")
sc <- score_method(X0, tf_targets, colnames(X0))
cat("TF-target pairs:", sc$n_pos, "pos /", sc$n_neg, "neg\n")
```

## Diagnostic: is this a smoothing-friendly (oracle t > 0) regime?

```{r oracle, message=FALSE}
e1 <- eval_all(X0, tf_targets, seed = 1L)
curve_df <- data.frame(t = as.integer(names(e1$curve)), AUC = e1$curve)
cat("AUC by global t (seed-101 panel):\n"); print(curve_df, row.names = FALSE)
cat(sprintf("oracle t = %d (AUC %.3f) | vanilla MCV t = %d | tolerant t = %d\n",
            e1$row$t_oracle, e1$row$auc_oracle, e1$row$t_vanilla, e1$row$t_tol))
```

## Multi-seed head-to-head: tolerant vs vanilla MCV vs rmagic_auto

```{r seeds, message=FALSE}
seeds <- 101:106
rows <- list()
for (s in seeds) {
  Xs <- build_panel(M, tf_targets, seed = s)
  r <- eval_all(Xs, tf_targets, seed = 1L)$row
  r$seed <- s; rows[[as.character(s)]] <- r
  cat(sprintf("seed %d | vanilla t=%d(%.3f) tol t=%d(%.3f) oracle t=%d(%.3f) rmagic=%.3f raw=%.3f\n",
              s, r$t_vanilla, r$auc_vanilla, r$t_tol, r$auc_tol,
              r$t_oracle, r$auc_oracle, r$auc_rmagic, r$auc_raw))
}
res <- do.call(rbind, rows)
write.csv(res, file.path(out_dir, "tusi_trajectory_results.csv"),
          row.names = FALSE)
```

```{r stats}
g_tv <- res$auc_tol - res$auc_vanilla
g_tr <- res$auc_tol - res$auc_rmagic
g_vr <- res$auc_vanilla - res$auc_rmagic
report <- function(x, lbl) {
  tt <- tryCatch(t.test(x), error = function(e) NULL)
  cat(sprintf("%-26s mean=%+.4f sd=%.4f  >0 in %d/%d%s\n", lbl, mean(x),
      sd(x), sum(x > 0), length(x),
      if (!is.null(tt) && !is.na(tt$p.value))
        sprintf("  (p=%.3f, 95%% CI [%+.4f,%+.4f])", tt$p.value,
                tt$conf.int[1], tt$conf.int[2]) else ""))
}
cat("Tusi 2018 basal BM trajectory,", length(seeds), "subsample seeds:\n")
report(g_tv, "tolerant - vanilla MCV")
report(g_tr, "tolerant - rmagic_auto")
report(g_vr, "vanilla  - rmagic_auto")
cat(sprintf("\nMedian t: vanilla=%g tolerant=%g oracle=%g\n",
            median(res$t_vanilla), median(res$t_tol), median(res$t_oracle)))
```

```{r fig, fig.width = 10, fig.height = 4}
p1 <- ggplot(curve_df, aes(t, AUC)) +
  geom_line(linewidth = .8, colour = "grey40") +
  geom_point(size = 2, colour = "grey40") +
  geom_point(data = subset(curve_df, t == e1$row$t_vanilla),
             aes(colour = "vanilla argmin"), size = 4) +
  geom_point(data = subset(curve_df, t == e1$row$t_tol),
             aes(colour = "tolerant"), size = 3) +
  geom_point(data = subset(curve_df, t == e1$row$t_oracle),
             aes(colour = "oracle"), size = 2) +
  scale_colour_manual(values = c("vanilla argmin" = "#1f77b4",
    "tolerant" = "#2ca02c", "oracle" = "#d62728"), name = NULL) +
  scale_x_continuous(breaks = 0:15) +
  labs(x = "global diffusion t", y = "TF-target AUC",
       title = "(a) Tusi inDrop: AUC-vs-t and selector landings") +
  theme_classic(base_size = 10) + theme(legend.position = "bottom")

pl <- res |> select(seed, auc_rmagic, auc_vanilla, auc_tol, auc_oracle) |>
  pivot_longer(-seed, names_to = "method", values_to = "AUC") |>
  mutate(method = recode(method, auc_rmagic = "rmagic_auto",
    auc_vanilla = "cv_mcv (vanilla)", auc_tol = "magic_tolerant",
    auc_oracle = "oracle"),
    method = factor(method, levels = c("rmagic_auto","cv_mcv (vanilla)",
      "magic_tolerant","oracle")))
p2 <- ggplot(pl, aes(method, AUC, fill = method)) +
  geom_boxplot(width = .5, outlier.shape = NA, alpha = .5) +
  geom_jitter(width = .12, size = 1.6, alpha = .8) +
  scale_fill_manual(values = c("rmagic_auto" = "#ff7f0e",
    "cv_mcv (vanilla)" = "#1f77b4", "magic_tolerant" = "#2ca02c",
    "oracle" = "#d62728"), guide = "none") +
  labs(x = NULL, y = "TF-target AUC",
       title = sprintf("(b) %d-seed robustness", length(seeds))) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

fig <- p1 + p2 + plot_layout(widths = c(1.15, 1))
ggsave(file.path(fig_dir, "fig_tusi_trajectory.png"), fig,
       width = 10, height = 4.2, dpi = 220, bg = "white")
fig
```
